---
title: Project Financial History
execute:
  echo: false
  date: today
---

```{r echo = F, message = F}

# clear environment
rm(list=ls())

# load packages
library(tidyverse)
library(readxl)
library(openxlsx)
library(data.table)
library(stringr)
library(magrittr)
library(janitor)
library(hms)
library(lubridate)
library(anytime)
library(xfun)
library(plotly)

```

*Updated `r Sys.Date()`, Benjamin Meyer, Kenai Watershed Forum (ben\@kenaiwatershed.org)*

# Summary

-   The Kenai River Baseline Water Quality Monitoring project is funded both direct as well as in-kind financial support from community partners.

-   Costs include contractual laboratory expenses, research staff personnel time, equipment, and other smaller expense categories. Partners are invoiced on an annual basis.

-   Since 2005, 56% inflation of the U.S. dollar has reduced the value of resources available to the project.

-   **In order to continue Kenai River water quality monitoring efforts, the project must either seek additional sources of funding or de-prioritize certain sampling sites or parameters.**

-   **Read the Partner Request Letter Here [Writing In Progress 10/28/2024]**

## Inflation and Declining Project Income

Since the year 2005, inflation of the US dollar has reduced it's value by 56%[^finance_history-1]. @fig-inflation-history plots this trend from 2005 - 2023.

[^finance_history-1]: <https://www.usinflationcalculator.com/>

<br>

```{r echo = F, message = F, warning = F}
# EXPENSES
# read in expense history data
expense_dat <- read_excel("other/input/financial/Agency Baseline_expenses timeline.xlsx",
                          sheet = "Sheet1", range = "D1:W32") %>%
  select(-TOTAL) %>%
  rename( "expense" = `...1`) %>%
  filter(expense != "Contractual",
         expense != "Contributions Income",
         expense != "Grants")

# group by expense categories
expense_dat_cat <- read.csv("other/input/financial/expense_categories_dat.csv") %>%
  remove_empty("cols")
expense_dat <- left_join(expense_dat,expense_dat_cat)
rm(expense_dat_cat)

# summarize & reformat
expense_dat_long <- expense_dat %>%
  pivot_longer(cols = "2005":"2022",names_to = "year") %>%
  group_by(expense_category,year)  %>%
  summarise("total_expenses" = sum(value)) 
write.csv(expense_dat_long,"other/temp.csv")

```

```{r echo = F, message = F, warning = F}
# INCOME
# read in and restructure income data to long format
income_dat <- read_excel("other/input/financial/Baseline Income History.xlsx", 
                  sheet = "income_history", range = "D4:W17") %>%
  rename("org" = "...1") %>%
  filter(!is.na(org)) %>%
  pivot_longer(cols = "2005":"2023",names_to = "year") %>%
  rename("income" = "value")
```

```{r echo = F, message = F, warning = F}
# INFLATION
# read in table of inflation history
inflation_dat <- read_excel("other/input/financial/usd_annual_inflation.xlsx") %>%
  select(year,inflation_value) %>%
  mutate(inflation_multiplier = 1-inflation_value,
         year = as.character(year)) %>%
  select(-inflation_value)

# join inflation history to income history
income_dat_annual <- left_join(income_dat,inflation_dat) %>%
  # calculate inflation adjusted income values
  mutate(inflation_adj_income = income*inflation_multiplier) %>%
  # summarise by year
  group_by(year) %>%
  summarise(income = sum(income, na.rm = T),
            inflation_adj_income = sum(inflation_adj_income, na.rm = T))

# join table of total annual expenses to income and inflation history
expense_dat_annual <- expense_dat_long %>%
  group_by(year) %>%
  summarise(total_expenses = sum(total_expenses))

annual_inf_exp_inc <- left_join(income_dat_annual,expense_dat_annual, by = "year") %>%
  filter(!is.na(total_expenses))

```

```{r echo = F, message = F, warning = F}
#| label: fig-inflation-history
#| fig-cap: "Inflation Adjusted Value of $1.00, Years 2005 - 2023"

# plot
inflation_dat %>%
  ggplot(aes(year,inflation_multiplier, group = inflation_multiplier)) +
  geom_point() +
  xlab("") +
  ylab("") +
  scale_y_continuous(labels = scales::dollar_format()) +
  ggtitle("Inflation Adjusted Value of $1.00, Years 2005 - 2023")



```

```{r echo = F}
# download inflation history
embed_file("other/input/financial/usd_annual_inflation.xlsx", text = "Download figure source data, Inflation History data of US Dollar 2005 - 2023")
```

<br>

Adjusted for inflation, project income has seen a steady decline from 2005 - 2022. @fig-inf-adj-inc displays project income history.

```{r echo = F, message = F, warning = F}
#| label: fig-inf-adj-inc
#| fig-cap: "Inflation Adjusted Project Income 2005 - 2022"

# plot
annual_inf_exp_inc %>%
  pivot_longer(cols = c("income","inflation_adj_income","total_expenses"), names_to = "category") %>%
  filter(category %in% c("inflation_adj_income")) %>%
  ggplot(aes(year,value, color = category, group = category)) +
  geom_line() +
  xlab("") +
  ylab("") +
  theme(legend.position = "none") +
  scale_y_continuous(labels = scales::dollar_format()) +
  ggtitle("Project Income, Inflation Adjusted Years 2005 - 2022")

```

## Income Detail

Project income history fluctuates annually as partner contributions vary. @fig-income-a displays annual cash contributions by individual partner organizations.

Note that in 2022, the project received a lump sum contribution from Kenaitze Indian Tribe that had been intended to be dispersed annually over years 2018 - 2022. Similar levels of annual contribution are not anticipated for future years. Additionally, in 2023 the project received a contribution from Salamatof Tribe via it's IGAP[^finance_history-2] program. As of 2024, the IGAP program is no longer active with Salamatof Tribe, thus further cash contributions are not anticipated.

[^finance_history-2]: <https://www.epa.gov/general-assistance-program-gap>

```{r echo = F, message = F, warning = F, fig.width = 12, fig.height = 12}
#| label: fig-income-a
#| fig-cap: "Income by partner organization"

# plot year vs amount by org, faceted bar chart
income_dat %>%
  ggplot(aes(x = year, y = income, label = income)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_grid(org ~ .) +
  theme(strip.text.y = element_text(angle = 360)) +
  scale_y_continuous(labels = scales::dollar_format()) +
  xlab("") +
  ylab("") +
  theme(plot.title = element_text(size = 16, face = "bold"),
        axis.text.y = element_text(size = 10),
        strip.text = element_text(size = 12)) +
  ggtitle("Project  Income History 2005-2023")

ggsave("other/output/financial/2005_2023_income_plot2.jpg", width = 10, height = 8)
```

With the atypical timing and receipt of funds from Kenaitze and Salamatof, the project saw increased income in 2022 and 2023; see @fig-income-b. However as seen in @fig-income-c, when adjusted for inflation, even these two high-income years are lower than typical total annual income received through year 2014.

```{r echo = F, message = F, warning = F, fig.height=6, fig.width=10}
#| label: fig-income-b
#| fig-cap: "Project income by partner organization"


# plot year vs income by org, stacked bar chart
income_dat %>%
  ggplot(aes(x = year, y = income, label = income, fill = org)) +
  geom_bar(stat = "identity") +
  scale_y_continuous(labels = scales::dollar_format()) +
  scale_fill_brewer(palette = "Paired", name = "Organization") +
  xlab("") +
  ylab("") +
  theme(plot.title = element_text(size = 16, face = "bold"),
        axis.text.y = element_text(size = 10)) +
  ggtitle("Project Income History 2005-2023")

ggsave("other/output/financial/2005_2023_income_plot1.jpg", width = 10, height = 8)

```


```{r echo = F, message = F,warning = F, fig.height=6, fig.width=10}
#| label: fig-income-c
#| fig-cap: "Project income by partner organization, adjusted for inflation relative from 2005 - 2023"

# read in annual inflation percentages, 2005 - 2023
# read in and join annual inflation multiplier values
inflation_dat <- read_excel("other/input/financial/usd_annual_inflation.xlsx") %>%
  select(year,inflation_value) %>%
  mutate(inflation_multiplier = 1-inflation_value,
         year = as.character(year)) %>%
  select(-inflation_value)

# join
income_dat <- left_join(income_dat,inflation_dat) %>%
  mutate(inflation_adj_value = income*inflation_multiplier)

# inflation adjusted plot 1
income_dat %>%
  ggplot(aes(x = year, y = inflation_adj_value, label = income, fill = org)) +
  geom_bar(stat = "identity") +
  scale_y_continuous(labels = scales::dollar_format()) +
  scale_fill_brewer(palette = "Paired", name = "Organization") +
  xlab("") +
  ylab("") +
  ggtitle("Inflation Adjusted Project Income History 2005-2023") +
  theme(plot.title = element_text(size = 16, face = "bold"),
        axis.text.y = element_text(size = 10))

ggsave("other/output/financial/2005_2023_inflation_adjusted_income_plot1.jpg", width = 10, height = 8)


```

# Expenses History

Project expense trends from years 2005 - 2022 are detailed in figure @fig-expense-history and source data is available in the link below the figure.

```{r echo = F, message = F, warning = F, fig.width = 14, fig.height = 10}
#| label: fig-expense-history
#| fig-cap: "Kenai River Baseline Water Quality Monitoring Expenses 2005 - 2022"

#plot 
  expense_dat_long %>%
  ggplot(aes(year,total_expenses, fill = expense_category)) +
  geom_bar(stat = "identity") +
  facet_grid(expense_category~.) +
  theme(strip.text.y = element_text(angle = 360),
        plot.title = element_text(size = 20, face = "bold"),
        axis.text.x = element_text(size =12, face = "bold"),
        axis.text.y = element_text(size = 14),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        strip.text = element_text(size = 14),
        legend.position = "none") +
  scale_y_continuous(labels = scales::dollar_format()) +
    labs(fill = "Expense\nCategory") + 
    ggtitle("Kenai River Baseline Water Quality Monitoring Expenses 2005 - 2022")

```

```{r echo = F}
# write csv
write.csv(expense_dat,"other/output/financial/expense_dat.csv", row.names = F)
embed_file("other/output/financial/expense_dat.csv", text = "Download Kenai River Baseline Water Quality Monitoring Project expense data 2005 - 2022")
```

In general, laboratory analysis expenses and staff scientist time are the largest expense categories. In years where little to no staff time was is recorded for the project, staff were funded by similar related projects.


## Source Data
All source data not linked at this page is available in the project GitHub repository at <https://github.com/Kenai-Watershed-Forum/kenai-river-wqx>
