# Regulatory Limits

We calculated regulatory limits based on freshwater quality standards listed from state and federal regulatory agencies...

## Metals

### Static Regulatory Threshold Values for Metals

Some regulatory threshold values for metals are static and do not depend on additional environmental parameters. The metals parameters in the downloadable table below have static thresholds assigned.

```{r echo = F, warning=FALSE, message=FALSE}

library(tidyverse)
library(janitor)
library(readxl)
library(xfun)
 
# read in reg threshold values for metals with static values
static_metals_reg_vals <- read_xlsx("other/input/regulatory_limits/master_reg_limits.xlsx", sheet = "static_regulatory_values") %>%
  filter(!is.na(agency),
         standard_type %in% c("drinking_water","irrigation_water","stock_water")) %>%
  remove_empty() %>%
  select(parameter_baseline_name,standard_type,reg_value,reg_unit) %>%
  pivot_wider(names_from = standard_type, values_from = reg_value) %>%
  rename(characteristic_name = parameter_baseline_name) %>%
  pivot_longer(cols = ends_with("water"), names_to = "standard", values_to = "value") %>%
  rename("Standard" = "standard")

# define directory
dir <- "other/input/regulatory_limits/formatted_reg_vals/static_metals_reg_vals.csv"

# export csv
write.csv(static_metals_reg_vals, dir, row.names = F)

# make viewable
embed_file(dir, text = "Download Regulatory Limits for Metals with Static Regulatory Threshold Values")

```

Static regulatory values are represented as horizontal lines in the following parameter-specific chapters.

<br>

### Metals with Hardness-Dependent Regulatory Values

Some regulatory values for metals are not simple static values, and are instead hardness-dependent. Hardness varies by site, by season, and with hydrologic conditions. We calculated hardness according to the following formula:

::: {.callout-note appearance="simple"}
Hardness = 2.497(Total Calcium in mg/L) + 4.119(Total Magnesium in mg/L)
:::

For sampling events with duplicate samples, we calculated average hardness.

```{r echo = F, warning=FALSE, message=FALSE}

# calculate hardness for all sample events in project history with available data

# source formatted data
dat <- read.csv("other/output/analysis_format/baseline_analysis_format.csv")

# calculate hardness column
hard <- dat %>%
  select(activity_identifier, activity_start_date, characteristic_name, result_measure_value) %>%
  filter(characteristic_name %in% (c("Magnesium","Calcium"))) %>%
  group_by(activity_identifier, activity_start_date, characteristic_name) %>%
  # get mean of duplicate samples
  summarise(avg_value = mean(result_measure_value)) %>%
  pivot_wider(names_from = "characteristic_name", values_from = "avg_value") %>%
  mutate(hardness = 2.497*(Calcium) + 4.119*(Magnesium)) %>%
  select(-Calcium, -Magnesium)

# join hardness back to overall dataset
dat <- left_join(dat,hard) 

# export hardness results
dir <- "other/output/hardness/dat_hardness.csv"
write.csv(dat, dir, row.names = F)

# make data downloadable
xfun::embed_file(dir, name = "Kenai River Baseline Hardness Data")

```

<br>

#### Acute and Chronic Hardness-Dependent Regulatory Threshold Values

We calculated hardness-dependent regulatory criteria based on formulas published on the Alaska Department of Environmental Conservation's Water Quality Standards Section website (<https://dec.alaska.gov/water/water-quality/standards/>). The "Alaska Water Quality Criteria Manual for Toxic and other Deleterious Organic and Inorganic Substances" (2022) document, along with the "Inorganic Toxics Criteria Worksheet" provide parameter-specific formulas for many substances.

Hardness-dependent criteria may be calculated from the following for freshwater metals:

::: {.callout-note appearance="simple"}
Acute (dissolved) = exp{mA\[ln(hardness)\]+ bA} (CF) Chronic (dissolved) = exp{mC \[ln(hardness)\] + bC} (CF)
:::

where mA, bA, mC, and bC are numerical parameters specific to each substance, and CF is the conversion factor for acute vs. chronic criteria.

These values are graphically displayed in the following chapters not juxtaposed against static horizontal lines, but instead they are highlighted by shape or color when the calculated threshold is exceeded.

```{r echo = F}

# hardness-dependent criteria
# see page 29 of "Alaska Water Quality Criteria Manual for Toxic and other Deleterious Organic and Inorganic Substances" for parameters and formulas

# read in parameters for hardness dependent criteria
diss_metals_hard_param <- read_excel("other/input/regulatory_limits/master_reg_limits.xlsx", skip = 1, sheet = "diss_metals_hard_parameters") %>%
  remove_empty() %>%
  select(-parameter_agency_name) %>%
  filter(!is.na(characteristic_name))

# join calculation parameters to overall dataset

hardness_dependent_metals_thresholds <- inner_join(dat, diss_metals_hard_param) %>%
  
  # Calculate hardness dependent criteria values
  
  ## ACUTE ##
  mutate(fw_acute_std = case_when(
    # cadmium
    characteristic_name == "Cadmium" ~ exp(m_a*log(hardness) + b_a) * (1.136672 - ((log(hardness)*0.041838))),
    # lead
    characteristic_name == "Lead" ~ exp(m_a*log(hardness) + b_a) * (1.46203 - ((log(hardness)*0.145712))),
    # all other acute hardness-dependent criteria
    .default = exp(m_a*log(hardness) + b_a)
  )) %>%
  
  ## CHRONIC ##
  mutate(fw_chronic_std = case_when(
    # cadmium
    characteristic_name == "Cadmium" ~ exp(m_c*log(hardness) + b_c) * (1.101672 - ((log(hardness)*0.041838))),
    # lead 
    characteristic_name == "Lead" ~ exp(m_c*log(hardness) + b_c) * (1.46203 - ((log(hardness)*0.145712))),
    # all other hardness-dependent criteria
    .default = exp(m_c*log(hardness) + b_c))) %>%
  
    # retain and rename needed columns
  select(characteristic_name, activity_start_date, result_sample_fraction_text, result_measure_value, result_measure_measure_unit_code, trib_mainstem, tributary_name, site_name, season, fw_acute_std, fw_chronic_std) %>%
  
  # calculate when exceedances occur
  ## acute
  mutate(fw_acute_exceed = case_when(
    result_measure_value >= fw_acute_std ~ "Y")) %>%
  
  ## chronic
  mutate(fw_chronic_exceed = case_when(
    result_measure_value >= fw_chronic_std ~ "Y")) 

# note: there is certainly a cleaner way to code the above, but for now am just reflecting the formula format as provided in the ADEC manual.


# define directory
dir <- "other/input/regulatory_limits/formatted_reg_vals/calculated_metals_reg_vals.csv"

# write csv of results
write.csv(hardness_dependent_metals_thresholds, dir, row.names = F)

# make csv downloadable
embed_file(dir, text = "Hardness Dependent Regulatory Thresholds for Metals - Kenai River Baseline Water Quality")

# it appears that export from EPA WQX of 2000 - 2014 data does not include status of samples as DUP or Trip Blank. Are results included at all? Also note that from Summer 2022 - present we are also using field blanks

# 11/16/23 working here to set up formulas for calculated criteria, see spreadsheet other/input/regulatory_limits

# 1.) read in static reg values
# 2.) calculate acute values
# 3.) calculate chronic values
# 4.) calculate other condition-dependent values (e.g. temp)
  
```

<br>

## Hydrocarbons

```{r echo = F}

### Diesel Range Organics 
### Gasoline Range Organics 
### Residual Range Organics 

# The above parameter names are included in the 2016 report. However it is unclear which specific parameters have equivalent names in the data downloaded from the EPA WQX

# Skip for now and return to as a later task

```

### Total BTEX

Total aromatic hydrocarbons (TAH) means the sum of the following volatile monoaromatic hydrocarbon compounds: benzene, toluene, ethylbenzene, and the xylene isomers, commonly called BTEX.

For each sampling event with available data, we calculated the sum of the parameters that compose BTEX and compared them to a static threshold value.

```{r echo = F}

# check
# does value for sum of "Benzene, toluene, ethyl benzene, xylenes mix" match value for sum of individual parameters? Are there events where both formats are reported?

```

<br>

## Nutrients (Nitrogen and Phosphorus)

```{r echo = F}

# The parameters in the existing dataset cover two names: 1.) "Inorganic nitrogen (nitrate and nitrite)" and 2.) "Nitrate"

# 2016 report: "inorganic nitrogen is generally present in well-aerated, natural streams in the form of nitrate (Glass, 1999)."

# just different names for the same parameter?

```

<br>

## Bacteria

```{r echo = F}

# page 15 18 AAC https://www.epa.gov/system/files/documents/2023-05/akwqs-chapter70-2023.pdf

# see description/code for beach project: https://dec.alaska.gov/media/23091/kenai-river-beach-2020-final-report.pdf



```

<br>

## Total Suspended Solids

```{r echo = F}

# pg 22 https://www.epa.gov/system/files/documents/2023-05/akwqs-chapter70-2023.pdf
```

<br>

## Other Parameters

```{r echo = F}

# pH, conductivity, temp, turb, DO

```
