# Regulatory Limits

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

```{r packages, echo = F, warning=FALSE, message=FALSE}

# clear environment
rm(list=ls())

# load packages
library(tidyverse)
library(janitor)
library(readxl)
library(xfun)
```

We calculated regulatory limits based on freshwater quality standards listed from state and/or federal regulatory agencies. Some regulatory threshold values are static and do not depend on additional environmental parameters, while some threshold values depend on local conditions like hardness, temperature, or pH.

We discuss how various parameters are assessed against standards below.

## Metals

### Metals with Static Regulatory Threshold Values

The metals parameters in the downloadable table below have *static* thresholds assigned.

[Static regulatory values are represented as solid horizontal lines in the chapters that follow.]{.underline} (to do: show graphic example of values above horizontal line here)

```{r echo = F}

# read in reg threshold values for metals with static values ###
static_metals_reg_vals <- read_xlsx("other/input/regulatory_limits/master_reg_limits.xlsx", sheet = "static_regulatory_values") %>%
  filter(!is.na(agency),
         standard_type %in% c("drinking_water","irrigation_water","stock_water")) %>%
  remove_empty() %>%
  select(parameter_baseline_name,standard_type,reg_value,reg_unit) %>%
  pivot_wider(names_from = standard_type, values_from = reg_value) %>%
  rename(characteristic_name = parameter_baseline_name) %>%
  pivot_longer(cols = ends_with("water"), names_to = "standard", values_to = "value") %>%
  rename("Standard" = "standard")

# define directory
dir <- "other/input/regulatory_limits/formatted_reg_vals/static_metals_reg_vals.csv"

# export csv
write.csv(static_metals_reg_vals, dir, row.names = F)

# make table viewable
embed_file(dir, text = "Download Regulatory Limits for Metals with Static Regulatory Threshold Values")
```

```{r}
# read in static reg values in wide format
static_metals_reg_vals <- read_xlsx("other/input/regulatory_limits/master_reg_limits.xlsx", sheet = "static_regulatory_values") %>%
  filter(static_category == "static_metals",
         standard_type %in% c("drinking_water","irrigation_water","stock_water")) %>%
  remove_empty() %>%
  select(parameter_baseline_name,standard_type,reg_value,reg_unit,agency,source) %>%
  pivot_wider(names_from = standard_type, values_from = reg_value) %>%
  rename(characteristic_name = parameter_baseline_name,
         reg_agency = agency,
         reg_agency_source = source) 

# source formatted data
dat <- read.csv("other/output/analysis_format/baseline_export_format.csv")
  

# join static regulatory values to parameter results, and identify when/where exceedences occur
static_metals_thresholds <- left_join(dat,static_metals_reg_vals) %>%
  mutate(drinking_water_exceed = case_when(
    drinking_water >= result_measure_value ~ "Y")) |>
  mutate(stock_water_exceed = case_when(
    stock_water >= result_measure_value ~ "Y")) |>
  mutate(irrigation_water_exceed = case_when(
    irrigation_water >= result_measure_value ~ "Y")) 

```

### Metals with Hardness-Dependent Regulatory Values

Some regulatory values for metals are not simple static values, and are instead hardness-dependent (i.e. toxicity depends on calcium (Ca) and magnesium (Mg) concentrations). The metals parameters in the downloadable table below have *hardness-dependent* thresholds assigned.
-   ~~**Add missing threshold values — Iron pipeline fix**~~ **DONE (3/31/2026).** Iron threshold added to `master_reg_limits.xlsx` and routed through the pipeline (see iron boxplot entry below). Note: Water Temperature and Fecal Coliform were also added directly to `all_reg_vals.csv` at this time, but that approach was superseded — direct edits to `all_reg_vals.csv` do not survive renders. **Correct fix applied 4/2/2026** — see April 2 session notes.

-   ~~**Update `threshold_table.R`**~~ **DONE (3/31/2026, superseded 4/2/2026).** Added `standard_labels`, `standard_authority`, and `unit_labels` entries for new standard type codes. **Superseded by April 2, 2026 change:** `threshold_table.R` now reads these mappings from the `standard_types` sheet in `master_reg_limits.xlsx` rather than hardcoded vectors.
# read in parameters which have hardness dependent criteria
diss_metals_hard_param <- read_excel("other/input/regulatory_limits/master_reg_limits.xlsx", skip = 1, sheet = "diss_metals_hard_parameters") %>%
  remove_empty() %>%
  select(-parameter_agency_name) %>%
  filter(!is.na(characteristic_name),
         !is.na(m_a))

# make downloadable list metals which have calculated parameters
dir <- "other/input/regulatory_limits/formatted_reg_vals/calculated_metals_list.csv"
write.csv(diss_metals_hard_param,dir, row.names = F)
embed_file(dir, text = "Download List of Metals with Hardness-Dependent Regulatory Threshold Values")

```

<br>

Hardness varies by site, by season, and with hydrologic conditions. We calculated hardness according to the following formula:

::: {.callout-note appearance="simple"}
Hardness = 2.497(Total Calcium in mg/L) + 4.119(Total Magnesium in mg/L)
:::

For sampling events with duplicate samples, we calculated average hardness. Note that some sampling events do not have calculated hardness values because total Ca and/or total Mg values were not available.

```{r echo = F, warning=FALSE, message=FALSE}

# calculate hardness for all sample events in project history with available Ca and Mg data

# confirm source for analysis-formatted data
dat <- read.csv("other/output/analysis_format/baseline_export_format.csv")

# calculate hardness column  

hard <- dat %>%
  select(activity_identifier, activity_start_date, characteristic_name, result_measure_value, result_measure_measure_unit_code,result_sample_fraction_text,monitoring_location_name) %>%
  filter(characteristic_name %in% (c("Magnesium","Calcium"))) %>%
  
  # retain only total Ca and Mg, not dissolved values
  filter(result_sample_fraction_text %in% (c("Total Recoverable","Unfiltered"))) %>%
  
  group_by(activity_start_date, monitoring_location_name, characteristic_name, result_measure_measure_unit_code) %>%
  
  # 2021 data has characteristic names in activity_identifier column fyi, observations prior to 2013 do not
  
  # get mean of all duplicate samples
  summarise(avg_value = mean(result_measure_value)) %>%
  pivot_wider(names_from = "characteristic_name", values_from = "avg_value") %>%
  # calculate hardness for each sample event
  mutate(hardness_mgl = 2.497*(Calcium) + 4.119*(Magnesium)) %>%
  ungroup() %>%
  select(-Calcium, -Magnesium,-result_measure_measure_unit_code)

# join hardness back to overall dataset
dat <- full_join(dat,hard,by = c("activity_start_date","monitoring_location_name")) 

# for the sampling events where hardness values are missing, why? presumably becuase total Ca and/or Mg values are missing. Confirm that this is the case. If "characteristic_name" does not include paired observations of Mg or Ca in 'hard_missing', than this is the case
hard_missing <- dat %>%
  filter(is.na(hardness_mgl)) %>%
  distinct(activity_identifier, .keep_all = T) %>%
  filter(characteristic_name %in% c("Calcium","Magnesium"))

# TO DO HERE: make sure that outlier/incorrect hardness values are not being applied in dissolved metals toxicity calculations


# spent inordinate time debugging the following join issue here: we can't include hardness units as a join column becuase some parameters have units other than mg/l

# export hardness joined to overall data results
dir <- "other/output/hardness/dat_hardness.csv"
write.csv(dat, dir, row.names = F)
### Completed this session (April 2, 2026)

-   ~~**Re-upload corrected `station.csv` to CDX (Task 1)**~~ **DONE.** Uploaded `other/output/wqx_formatted/station.csv` with `HorizontalCollectionMethodName = "GPS-Unspecified"` and `HUCEightDigitCode = 19020302` for all 22 stations. No import configuration changes were needed (the existing A–K column mapping was correct). **Verify in WQP after 24–72 hours** that HUC codes are now populated for all 22 KBL stations, then check HMW tributary visibility.

    **Pipeline durability confirmed:** the `GPS-Unspecified` fix is hardcoded in `appendix_a.qmd`'s station generation block and applies on every render. The `case_when` now catches both `"Interpolation-Satellite"` and `"Unknown"` as a safety net. `WaterbodyName` was investigated as an additional improvement but is **not available as an importable field** in the WQX Web CSV batch import UI — it does not appear in the Elements list. The field can only be set manually via the WQX Web Review interface on individual station records.

-   ~~**Move `standard_labels` / `standard_authority` from hardcoded R to `master_reg_limits.xlsx`**~~ **DONE.** Added a `standard_types` sheet to `master_reg_limits.xlsx` with columns: `standard_type` (code), `display_label`, `regulatory_authority`, `review_needed`. Contains 20 rows covering all standard type codes from both the spreadsheet and `threshold_table.R`. Updated `threshold_table.R` to read from this sheet via `readxl::read_excel()` instead of hardcoded named vectors. The `pick_list` sheet is now superseded for this purpose. **Five rows flagged `review_needed = Y`** (inferred labels/authorities for codes not previously in `threshold_table.R`): `fw_acute`, `fw_chronic`, `harvest_aquatic_life`, `noncarc_aquatic_org`, `noncarc_water`, `secondary_water_recreation` — verify against 18 AAC 70 and USEPA criteria docs before final render.

-   ~~**Fix Water Temperature and Fecal Coliform threshold values so they survive renders**~~ **DONE.** Root cause: these values had been added directly to `all_reg_vals.csv`, but that file is regenerated on every render from `master_reg_limits.xlsx` via `static_boxplot_function.R`, so direct edits were silently lost. Fix applied in three places (matching the Iron pipeline pattern):

    1.  **`master_reg_limits.xlsx` → `static_regulatory_values` sheet** — 5 new rows with `static_category = "field_bio_standards"`: Water Temp (20/15/13°C, `deg_c`) and Fecal Coliform (400/40 CFU/100mL, `cfu/100ml`), all sourced to ADEC 18 AAC 70.020.
    2.  **`reg_limits.qmd`** — new export block after the Iron block: filters `static_category == "field_bio_standards"`, writes `other/input/regulatory_limits/formatted_reg_vals/field_bio_reg_vals.csv`.
    3.  **`functions/static_boxplot_function.R`** — reads `field_bio_reg_vals.csv` and includes it in the `bind_rows()` call.

-   ~~**Revise boxplot legend for clarity and correctness**~~ **DONE.** Both the ggplot2 (DOCX) and plotly (HTML) legends now reflect the distinction between static and hardness-dependent thresholds. Changes to `functions/static_boxplot_function.R`:

    -   **`std_labels` moved to top level** of the script (computed once on source, shared by `create_facet_plots()` and `clean_plotly_legend()`). Reads from the `standard_types` sheet in `master_reg_limits.xlsx`; labels are `str_wrap()`'d at 20 characters.
    -   **`linetype_map` expanded** to cover all 20 known standard type codes (previously only 8 were defined; new codes for FC, Water Temp, Iron, etc. would fall back to default linetypes and raw code strings).
    -   **`scale_linetype_manual()` now uses `labels = std_labels[present_standards]` and `breaks = present_standards`**, so only standards actually present for the current parameter appear in the legend, with human-readable wrapped labels.
    -   **Legend sub-titles:** `scale_linetype_manual(name = "Static regulatory\nthresholds")` and `scale_color/shape_manual(name = "Hardness-dependent\nexceedance type")`. Guides are ordered (static thresholds first) and stacked vertically inside a single bordered box.
    -   **`clean_plotly_legend()` extended** to handle both ggplotly trace naming patterns (facet-suffix `(NAME,N)` and bare names, which occur when no color/shape aesthetic is mapped). Traces are classified into `legendgroup = "threshold"` or `"exceedance"`. Per-group title text (`legendgrouptitle`) was removed — it caused duplicate label rendering in the HTML output. Instead, a single "Legend" title is applied at the layout level. All layout properties (title, bgcolor, bordercolor, borderwidth, `tracegroupgap = 20`) are set via **direct property assignment** rather than `plotly::layout()`, because `layout()` does not reliably override properties already set by `ggplotly()`.

### Tasks for next session (in order)

1.  ~~**Verify corrected `station.csv` in WQP and HMW.**~~ **DONE (April 2, 2026).** All 22 tributary and mainstem KWF sites are now visible in HMW with their associated monitoring data. GPS-Unspecified fix confirmed working.

    **Two HMW data display issues identified (April 2, 2026) - see Tasks 1a and 1b below.**

1a. **\[HIGH PRIORITY\] Standardize sample fraction names across all years for HMW display.** HMW displays data filtered by sample fraction and cannot simultaneously show records with different fraction names that represent the same thing. Known inconsistency: dissolved metals (e.g., Copper, Zinc) appear as both `"Filtered, field"` and `"Dissolved"` across different data years — HMW can only display one at a time, splitting the record of the same parameter across two views. **Action items:** (a) audit all distinct `ResultSampleFractionText` values in WQP across all KWF years using `dataRetrieval`; (b) identify all cases where multiple fraction names describe the same physical sample type; (c) determine the correct WQX domain value for each (WQX domain list is authoritative); (d) implement a normalization step in `appendix_a.qmd`'s CDX export block that maps all variants to the canonical value; (e) apply the same normalization in the annual QA/QC pipeline in `kenai-river-wqx-qaqc`; (f) generate corrected CDX upload files for all historical years with non-canonical fraction names and re-upload. This will require a full CDX re-upload for affected years.

1b. **\[HIGH PRIORITY\] Standardize characteristic (parameter) names across all years for HMW display.** HMW treats differently-named characteristics as separate parameters, preventing unified display. Known inconsistency: Nitrate + Nitrite data exists under at least three names across KWF's historical and 2021 records: `"Nitrate + Nitrite"`, `"Nitrite"` (likely mislabeled), and `"Inorganic nitrogen (nitrate and nitrite) ***retired***use Nitrate + Nitrite"` (a retired WQX domain value). These appear to all represent the same analyte (method 4500-NO3(F) measures combined nitrate + nitrite). **Action items:** (a) audit all distinct `CharacteristicName` values in WQP across all KWF years; (b) cross-reference against current WQX domain list to identify retired or non-canonical names; (c) map all variants to the current canonical WQX characteristic name; (d) implement normalization in the CDX export block and annual pipeline; (e) re-upload corrected files for affected historical years. **Note:** changing a `CharacteristicName` in WQX may change the `ActivityID` for associated records — verify whether a delete + re-upload is required rather than a simple update.

2.  **\[HIGH PRIORITY\] Fix HMW visibility for 15 legacy numeric-ID stations - move process to `kenai-river-wqx-qaqc` repo and carry out the upload.** During session 3/31/2026, a corrective station upload file was generated at `other/output/epa_wqp_uploads/corrected_epa_wqp_uploads/station_historical_correction.csv` for 15 historical stations (IDs `KENAI_WQX-10000063` through `KENAI_WQX-10000131`, named `KBX_m_*`, covering 2002-2009, 3,884 results). These stations have `HorizontalCollectionMethodName = "Unknown"` and blank `HUCEightDigitCode` in WQP. **This process belongs in `kenai-river-wqx-qaqc`, not here.** Action items: (a) recreate the corrective upload script in `kenai-river-wqx-qaqc`; (b) upload to CDX; (c) verify HUC codes in WQP; (d) delete `station_historical_correction.csv` from this repo.

2a. **HMW tributary visibility - ADEC liaison.** Contact ADEC liaison to request that KWF tributaries be added to a future Integrated Report assessment cycle — KWF's WQP dataset is the primary data source for these streams, and assessed waterbody status in ATTAINS is required for tributaries to show condition status in HMW (not just monitoring data). Note: `WaterbodyName` cannot be batch-imported via the WQX Web CSV import UI (field not available in the Elements list); set manually via WQX Web Review if desired.

3.  **Verify boxplot DOCX sizing fix.** Render DOCX and confirm that jitter points, facet strip text, and legend text are visibly larger than in HTML output. The fix is in `functions/static_boxplot_function.R` - the `is_docx` detection uses `Sys.getenv("QUARTO_PROFILE") == "docx"` as the primary check. If still not working, add `cat(Sys.getenv("QUARTO_PROFILE"), "\n")` in a temporary test chunk to confirm the environment variable is set during rendering.

4.  **Verify `review_needed = Y` rows in the `standard_types` sheet of `master_reg_limits.xlsx`.** Six standard type codes have inferred labels/authorities: `fw_acute`, `fw_chronic`, `harvest_aquatic_life`, `noncarc_aquatic_org`, `noncarc_water`, `secondary_water_recreation`. Confirm display labels and regulatory authority against 18 AAC 70 and USEPA criteria documents before final render. Once confirmed, set `review_needed = N` for each.

#### Acute and Chronic Hardness-Dependent Regulatory Threshold Values

Separate hardness-dependent standards often apply for acute vs. chronic metals toxicity.

We calculated hardness-dependent regulatory criteria based on formulas published on the Alaska Department of Environmental Conservation's Water Quality Standards Section website (<https://dec.alaska.gov/water/water-quality/standards/>). The "Alaska Water Quality Criteria Manual for Toxic and other Deleterious Organic and Inorganic Substances" (2022) document, along with the "Inorganic Toxics Criteria Worksheet" provide parameter-specific formulas for many substances.

Hardness-dependent criteria is calculated from the following formula for metals in freshwater:

::: {.callout-note appearance="simple"}
Acute (dissolved) = exp{mA\[ln(hardness)\]+ bA} (CF) Chronic (dissolved) = exp{mC \[ln(hardness)\] + bC} (CF)
:::

where mA, bA, mC, and bC are numerical parameters specific to each substance, and CF is the conversion factor for acute vs. chronic criteria.

These values are graphically displayed in the following chapters. Rather than juxtaposed against static horizontal lines, instead they are highlighted by shape or color when the calculated threshold is exceeded; specific to that sampling event's hardness conditions. See **figure X** for an example

**(to do: show graphic example of values above threshold as colored asterisks here)**

```{r echo = F}

# define metals with hardness dependent criteria
metals_hcrit <- c("Cadmium","Chromium","Copper","Lead","Nickel","Silver","Zinc")

# reduce overall dataset to just metals with hardness-dependent parameters
dat_metals_hard_dependent <- dat %>%
  filter(characteristic_name %in% metals_hcrit)

# join hardness-dependent calculation parameters to overall dataset
hardness_dependent_metals_thresholds <- full_join(dat_metals_hard_dependent, diss_metals_hard_param, by = "characteristic_name") %>%

  # Calculate hardness dependent criteria values
  
  ## ACUTE ##
  mutate(fw_acute_std = case_when(
    # cadmium
    characteristic_name == "Cadmium" ~ exp(m_a*log(hardness_mgl) + b_a) * (1.136672 - ((log(hardness_mgl)*0.041838))),
    # lead
    characteristic_name == "Lead" ~ exp(m_a*log(hardness_mgl) + b_a) * (1.46203 - ((log(hardness_mgl)*0.145712))),
    # all other acute hardness-dependent criteria
    .default = exp(m_a*log(hardness_mgl) + b_a)
  )) %>%
  
  ## CHRONIC ##
  mutate(fw_chronic_std = case_when(
    # cadmium
    characteristic_name == "Cadmium" ~ exp(m_c*log(hardness_mgl) + b_c) * (1.101672 - ((log(hardness_mgl)*0.041838))),
    # lead 
    characteristic_name == "Lead" ~ exp(m_c*log(hardness_mgl) + b_c) * (1.46203 - ((log(hardness_mgl)*0.145712))),
    # all other hardness-dependent criteria
    .default = exp(m_c*log(hardness_mgl) + b_c))) %>%
  
# retain and rename needed columns
# select(characteristic_name, activity_start_date,result_sample_fraction_text,result_measure_value,result_measure_measure_unit_code)

#trib_mainstem, tributary_name, site_name, season, fw_acute_std, fw_chronic_std) %>%
  
  
# note that all dissolved metals units are in ug/l, and all hardness units are in mg/l; as provided in the ADEC spreadsheet referenced earlier in this document (toxicsbook-adec-wqstandards-01-11-24)
  
  # calculate when exceedences occur
  ## acute
  mutate(fw_acute_exceed = case_when(
    result_measure_value >= fw_acute_std ~ "Y")) %>%
  
  ## chronic
  mutate(fw_chronic_exceed = case_when(
    result_measure_value >= fw_chronic_std ~ "Y")) 

# note: there is certainly a cleaner way to code the above, but for now am just reflecting the formula format as provided in the ADEC manual.


# define directory
dir <- "other/input/regulatory_limits/formatted_reg_vals/calculated_metals_reg_vals.csv"

# write csv of results
write.csv(hardness_dependent_metals_thresholds, dir, row.names = F)

# make csv downloadable
embed_file(dir, text = "Hardness Dependent Regulatory Thresholds for Metals - Kenai River Baseline Water Quality")

# it appears that export from EPA WQX of 2000 - 2014 data does not include status of samples as DUP or Trip Blank. Are results included at all? Also note that from Summer 2022 - present we are also using field blanks

# join hardness info and calculations and results back to verall dataframe
dat <- left_join(dat,hardness_dependent_metals_thresholds)


# note: the data visualization boxpot function filters out any "rejected" results
```

```{r}
# possible general approach for graphic plots: export sheets of static and hardness-dependent results; use this to intake for all figures
```

<br>

## Hydrocarbons

Note: the 2016 baseline report presents hydrocarbon data chapters structured as follows:

-   Diesel Range Organics

-   Gasoline Range Organics

-   Residual Range Organics

However, as of `r Sys.Date()` it is unclear which specific hydrocarbon parameters, or grouped sums of hydrocarbon parameters, have equivalent names in our data set downloaded from the EPA WQX. As a result, in this report we will structure data to simply assess BTEX values against the ADEC regulatory standard. There may be federal standards that also apply to hydrocarbons, to be addressed later.

```{r echo = F}

# prepare static threshold reg value for BTEX
hydrocarbon_reg_vals <- read_xlsx("other/input/regulatory_limits/master_reg_limits.xlsx", sheet = "static_regulatory_values") %>%
  filter(static_category == "hydrocarbons") %>%
  remove_empty() %>%
  select(parameter_baseline_name,standard_type,reg_value,reg_unit) %>%
  pivot_wider(names_from = standard_type, values_from = reg_value) %>%
  rename(characteristic_name = parameter_baseline_name) %>%
  pivot_longer(cols = ends_with("water"), names_to = "standard", values_to = "value") %>%
  rename("Standard" = "standard") %>%
  filter(!is.na(value)) %>%
  select(-"NA")

# define directory for where hydrocarbon reg values are saved
dir <- "other/input/regulatory_limits/formatted_reg_vals/hydrocarbon_reg_vals.csv"

# export csv
write.csv(hydrocarbon_reg_vals, dir, row.names = F)

# make table viewable
embed_file(dir, text = "Download Regulatory Limits for Hydrocarbons with Static Regulatory Threshold Values")

```

### Total BTEX

Total aromatic hydrocarbons (TAH) means the sum of the following volatile monoaromatic hydrocarbon compounds: benzene, toluene, ethylbenzene, and the xylene isomers, commonly called BTEX. The Alaska Dept of Environmental Conversation describes a maximum static regulatory value of 10 ug/l BTEX.

For each sampling event with available hydrocarbons data, we calculated the sum of the parameters that compose BTEX and compared them to a static threshold value.

To prepare our existing data to be assessed against this standard, we addressed the following issues:

-   In some cases, BTEX data is quantified as a total mixture (sum of all benzene, toluene, ethylbenzene, and the xylene isomers). In other cases, the individual substances are instead quantified separately.

-   In yet other cases, both the values of individual components of BTEX and the total mixture values are reported separately.

-   To create a consistent parameter, we summed the total quantities of benzene, toluene, ethylbenzene, and xylene by unique sampling event (unique location/time) when it had not already been reported as such.

-   In some cases, multiple isomers of the same hydrocarbon are quantified (e.g. o-xylene vs m,p-xylene). In these cases, we summed the total quantity of the substance for all isomers.

```{r echo = F}

# Goal here: replace old volatiles data in the overall dataframe with the aggregagated btex data we discussed above. 

# specify isomers to be aggregaed into single compound
xylene_isomers <- c("m,p-Xylene" = "Xylene","o-Xylene" = "Xylene")

# summarise btex characteristic_name because it is long and cumbsersome

# 1.) aggregate all volatiles data to sum into "`Benzene, toluene, ethyl benzene, xylenes mix"
btex_dat <- dat %>%
| `functions/static_boxplot_function.R` | Builds `plots` list (tributary + river mile faceted boxplots); defines `clean_plotly_legend()`. Reads `std_labels` from `master_reg_limits.xlsx` → `standard_types` sheet at top level (shared by both functions). |
| `functions/render_plots.R` | Defines `render_parameter_plots(plots)`: returns `htmltools::tagList` for HTML, calls `print()` for DOCX |
| `functions/threshold_table.R` | Defines `show_threshold_table(characteristic, no_threshold_note = NULL)`. Reads display labels and regulatory authority from `master_reg_limits.xlsx` → `standard_types` sheet. |
  # round all values to two decimals
  mutate(result_measure_value = round(result_measure_value,2)) %>%
    # if duplicate samples exist, take their average
  group_by(activity_identifier,characteristic_name) %>%
  summarise(result_measure_value = mean(result_measure_value)) %>%
   # restructure dataframe
  pivot_wider(names_from = "characteristic_name", values_from = "result_measure_value") %>%
  mutate(btex_sum = sum(Benzene,Ethylbenzene,Toluene,Xylene,`o-Xylene`,`m,p-Xylene`, na.rm = T)) %>%
  # does the reported sum match the calculated sum?
  mutate(match = case_when(
    as.character(`Benzene, toluene, ethyl benzene, xylenes mix`) == as.character(btex_sum) ~ "Y",
    TRUE ~ "N")) %>%
  # for instances where no provided sum of volatiles exists, use the calculated value
  mutate(`Benzene, toluene, ethyl benzene, xylenes mix` = case_when(
    is.na(`Benzene, toluene, ethyl benzene, xylenes mix`) ~ sum(Benzene,Ethylbenzene,Toluene,Xylene,`o-Xylene`,`m,p-Xylene`, na.rm = T),
    TRUE ~ `Benzene, toluene, ethyl benzene, xylenes mix`
  )) %>%
  select(activity_identifier,`Benzene, toluene, ethyl benzene, xylenes mix`) %>%
  rename(MEAN_result_measure_value = `Benzene, toluene, ethyl benzene, xylenes mix`) %>%
  mutate(characteristic_name = "Benzene, toluene, ethyl benzene, xylenes mix")


# 2.) extract a seperate dataframe of only volatiles from overall dataset
all_volatiles <- dat %>%
  filter(result_sample_fraction_text == "Volatile")

# 3.) Use this extracted dataset to match up with the aggregated dataset of volatiles only
mean_volatiles <- left_join(all_volatiles,btex_dat, by = c("activity_identifier","characteristic_name")) %>%
  # keep only the values from step 1
  filter(!is.na(MEAN_result_measure_value)) %>%
  select(-result_measure_value) %>%
  rename(result_measure_value = MEAN_result_measure_value)

# 4.) Use complete aggregated volatiles dataset to replace other volatiles info in overall dataset
dat <- dat %>%
  filter(is.na(result_sample_fraction_text) | result_sample_fraction_text != "Volatile") %>%
  # bind aggregated volatiles values back to verall dataset
  bind_rows(mean_volatiles)
  
# types of issues we addressed in preparing BTEX data
# - value exists only for total btex sum column (we left these as is)
# - no provided sum value exists, only calculated (we used the calculated value)

# hydrocarbons data should now be prepared in a format it can be plotted and assessed
```

<br>

## Nutrients (Nitrogen and Phosphorus)

```{r echo = F}
# TO DO - combine nitrogen names into single parameter!

# exploring nutrients data structure

# specify nutrients parameter names
nutrients_parameters <- c("Phosphorus","Nitrate + Nitrite")

# create nutrients dataframe
nutrients <- dat %>%
  filter(characteristic_name %in% nutrients_parameters)

# nitrogen only
nitrogen <- nutrients %>%
## Regulatory Limits

-   Standards are Alaska and Federal freshwater aquatic life criteria
-   Some metal thresholds (Cadmium, Chromium, Copper, Lead, Zinc) are **hardness-dependent** - threshold values are calculated per-sample based on measured hardness
-   Regulatory values are stored in `output/regulatory_values/`
-   Session variables `static_metals_thresholds`, `static_metals_reg_vals`, `diss_metals_hard_param` hold processed threshold data

### Regulatory Threshold Architecture

`master_reg_limits.xlsx` is the **single source of truth** for all regulatory threshold data. Sheets:

| Sheet | Contents |
|---|---|
| `static_regulatory_values` | All static threshold values by parameter and standard type. Categories: `static_metals`, `hydrocarbons`, `nutrients`, `other`, `total_metals_aquatic_life` (Iron), `field_bio_standards` (Water Temp, Fecal Coliform). |
| `calculated_regulatory_values` | Hardness-dependent threshold formulas (Cd, Cr, Cu, Pb, Zn). |
| `diss_metals_hard_parameters` | Parameters for hardness-dependent calculations. |
| `standard_types` | Display labels and regulatory authority for all standard type codes. Used by `threshold_table.R` and `static_boxplot_function.R`. Six rows flagged `review_needed = Y` — verify before final render. |
| `pick_list` | Legacy list of standard type codes — superseded by `standard_types` sheet. |

**Adding a new static threshold:** add a row to `static_regulatory_values` with the appropriate `static_category`, add or confirm the `standard_types` entry, add an export block in `reg_limits.qmd` if a new category is needed, and add the resulting CSV to the `bind_rows()` in `static_boxplot_function.R`. See the Iron and field_bio_standards entries as examples.
  as.character()

# how many observations of "Nitrate + Nitrite"
inorganic_n_count <- dat %>%
  filter(characteristic_name == "Nitrate + Nitrite") %>%
  count() %>%
  as.character()

# make all N observations in to just "Nitrate" for characteristics name; see discussion below
dat <- dat %>%
  mutate(characteristic_name = case_when(
    characteristic_name == "Nitrate" ~ "Nitrate + Nitrite",
    TRUE ~ characteristic_name))


# note: there appears to be one instance where the analytical method for nitrogen is reported as 200.7(W). Obviously an error, fix it.

```

The following nurtient parameters names are present in the dataset:

-   **Output CSVs regenerated (March 27, 2026, clean).** Both `other/output/wqx_formatted/2021_kwf_baseline_results_wqx.csv` and `2021_export_data_flagged.csv` have been regenerated with corrected Ca/Mg/Fe values and correct RM 1.5 site IDs. **2021 CDX re-upload completed 3/31/2026** (all three files: `results_activities.csv`, `project.csv`, `station.csv`). WQP download on 3/31/2026 confirms 835 records present. **Corrected `station.csv` re-uploaded 4/2/2026** with `HorizontalCollectionMethodName = "GPS-Unspecified"` and `HUCEightDigitCode = 19020302` for all 22 stations. Verify in WQP after 24–72 hours propagation (see Task 1).
-   "Inorganic nitrogen (nitrate and nitrite)"
-   **Hydrocarbon data** missing from 2025 WQP download (uploaded Jan 2024 but not appearing in download).

-   **Sample fraction name inconsistency across years (identified April 2, 2026).** Dissolved metals (e.g., Copper, Zinc) appear in WQP under both `"Filtered, field"` and `"Dissolved"` depending on the data year. HMW filters by fraction and cannot display both simultaneously, splitting what should be a unified parameter record. Needs audit and corrective CDX upload for affected years. See Task 1a.

-   **Characteristic name inconsistency across years (identified April 2, 2026).** Nitrate + Nitrite data exists in WQP under at least three names: `"Nitrate + Nitrite"`, `"Nitrite"`, and `"Inorganic nitrogen (nitrate and nitrite) ***retired***use Nitrate + Nitrite"`. These are all believed to represent the same analyte (method 4500-NO3(F)). The retired name was a WQX domain value that has since been superseded. Needs audit of all characteristic names across all KWF years against the current WQX domain list, followed by corrective CDX re-upload. See Task 1b.

The previous reports on these data (Guerron Orejuela 2016 and McCard 2007) use only the terms "Nitrate" and "Phosphorus" to describe nutrients.

In our dataset, we have `r nitrate_count` observations of "Nitrate" and `r inorganic_n_count` observations of "Inorganic nitrogen (nitrate and nitrite)." Both of these parameters are reported to use the identical lab analytical method (4500-NO3(E)). For this report, will treat the two parameter names as indistinguishable, and data appears in the figure as such.

<br>

## Fecal Coliform Bacteria

A variety of fecal coliform standards are described in ADEC standards ([18 AAC 70](https://dec.alaska.gov/water/water-quality/standards/)). Fecal coliform bacteria may be assessed against criteria in two ways:

-   Instantaneous point values, where criteria are exceeded if 10% of samples are above threshold within a 30 day period
-   Geometic mean within a 30-day period

In this project, we collect only single samples on single days for each site. Thus, data are not structured to be assessed against existing state standards. In this report, we display two fixed threshold values for a) harvesting aquatic life and b) secondary recreation (contact with water), but it should be notes that single values above these lines do not represent regulatory exceedences.

-   helpful data to understand where/when more monitoring is needed
-   important to help understand spatial trends

Additional sampling beyond this project is needed to reasonably characterize bacteria conditions in the Kenai River, and several major efforts between 2010 - 2020 contributed to this effort. For more information and detailed reports on Kenai River bacteria monitoring, visit <https://www.kenaiwatershed.org/other-water-research/> and click on "Bacteria Monitoring."

```{r echo = F}


# page 15 18 AAC https://www.epa.gov/system/files/documents/2023-05/akwqs-chapter70-2023.pdf

# see description/code for beach project: https://dec.alaska.gov/media/23091/kenai-river-beach-2020-final-report.pdf

# the horizontal lines are not appearing in the fecal coliform chapter for some reason, diagnose

```

<br>

## Total Suspended Solids

```{r echo = F}
# pg 22 https://www.epa.gov/system/files/documents/2023-05/akwqs-chapter70-2023.pdf


```

Alaska Department of Environmental Conservation (ADEC) standards require that fine sediment in the range of 0.1 mm to 4.0 mm cannot increase over 5% by weight higher than natural conditions in gravel beds used by anadromous or resident fish for spawning, and the sediment can never exceed a maximum of 30% by weight in freshwater for the growth and propagation of fish, shellfish, other aquatic life, and wildlife [@adec2025]. More generally, the ADEC standard also states that the amount of deposited or suspended sediment cannot cause adverse effects on aquatic life, including their habitat and reproduction.

As such, there is no fixed standard for total suspended solids. Data from this project is best interpreted by either examining long term trends and/or evaluating results in the context of data that provide evidence for "natural conditions."

<br>

## Other Parameters

```{r echo = F}

# pH
# conductivity: no fixed standard
# temp: replace with discussion/reference to AKTEMP and modern literature
# turbidity: no fixed standard
# DO: not analysed in 2016 report

# following the template flow below , read in parameter thresholds FROM master sheet and format them to match existing column names (characteristic_name,	reg_unit,	Standard,	value)
# once there, join thresholds std table to reg_vals table in static plot fxn


# read in pH threshold values
ph_reg_vals <- read_excel("other/input/regulatory_limits/master_reg_limits.xlsx", sheet = "static_regulatory_values") %>%
  remove_empty() %>%
  select(-parameter_agency_name) %>%
  # select for pH information only
  filter(parameter_baseline_name == "pH") %>%
  select(-agency,-source,-special_note,-data_entry) %>%
  rename("characteristic_name" = "parameter_baseline_name",
         "value" = "reg_value",
         "Standard" = "standard_type")

# make downloadable list of pH parameter standards 
dir <- "other/input/regulatory_limits/formatted_reg_vals/ph_reg_vals.csv"
write.csv(ph_reg_vals,dir, row.names = F)
embed_file(dir, text = "Download List of pH Regulatory Threshold Values")

# Total metals with freshwater aquatic life standards (e.g. Iron fw_chronic = 1 mg/L)
# Note: these values are stored in mg/L (not ug/L) to match analysis data units.
total_metals_aq_reg_vals <- read_excel("other/input/regulatory_limits/master_reg_limits.xlsx",
                                       sheet = "static_regulatory_values") %>%
  remove_empty() %>%
  filter(static_category == "total_metals_aquatic_life") %>%
  select(-parameter_agency_name, -agency, -source, -special_note, -data_entry) %>%
  rename("characteristic_name" = "parameter_baseline_name",
         "value"               = "reg_value",
         "Standard"            = "standard_type")

dir <- "other/input/regulatory_limits/formatted_reg_vals/total_metals_aq_reg_vals.csv"
write.csv(total_metals_aq_reg_vals, dir, row.names = FALSE)

# Field and biological parameters with static thresholds (Water Temperature, Fecal Coliform)
field_bio_reg_vals <- read_excel("other/input/regulatory_limits/master_reg_limits.xlsx",
                                 sheet = "static_regulatory_values") %>%
  remove_empty() %>%
  filter(static_category == "field_bio_standards") %>%
  select(-parameter_agency_name, -agency, -source, -special_note, -data_entry) %>%
  rename("characteristic_name" = "parameter_baseline_name",
         "value"               = "reg_value",
         "Standard"            = "standard_type")

dir <- "other/input/regulatory_limits/formatted_reg_vals/field_bio_reg_vals.csv"
write.csv(field_bio_reg_vals, dir, row.names = FALSE)

```

<br>

The following parameters are also monitored for each sampling event:

-   pH - The ADEC standard for pH maintains that freshwater may not have a pH less than 6.5 or greater than 8.5 for the growth and propagation of fish, shellfish, other aquatic life and wildlife.

-   Water Temperature - See [https://aktemp.uaa.alaska.edu/](https://aktemp.uaa.alaska.edu/#/)

-   Turbidity - see chapter on turbidity for details

-   Dissolved Oxygen - monitored, but not reported in 2016 study

```{r echo = F, message = F}
# end of script task: combine dataframes from all parameters evaluated against regulatory thresholds. export into "other/output/analysis_format", then source this table for the boxplot and table download functions

# Metals with Static Regulatory Threshold Values
## static_metals_thresholds

# Metals with Hardness-Dependent Regulatory Threshold Values
## hardness_dependent_metals_thresholds

# Hydrocarbons
## 

# Nutrients

# Bacteria

# Total suspended solids

# Field parameters

# working here 3/28/25 to ensure what's spit out of here still retains the result_status_identifier column

# this data below has been modified as described above to ensure compatibility between parameter dimensions and evaluation standards
write.csv(dat,"other/output/analysis_format/baseline_analysis_format.csv", row.names = F)

```
