---
execute:
  echo: false
  cache: false
date: "`r Sys.Date()`"
format:
  html:
    code-fold: true
    code-tools: true
    code-summary: "Show code"
---

# Annual QA/QC Pipeline: {YEAR} Data

## Introduction

This document contains the complete data ingestion, QA/QC evaluation, flagging, and CDX export workflow for the {YEAR} Kenai River Baseline Water Quality Monitoring dataset.

The structure of this document follows the Data Evaluation Checklist provided by the Alaska Department of Environmental Conservation. Field observations that do not meet QA/QC standards are flagged before upload to the EPA Water Quality Exchange (WQX).

For the 2021 worked example (with full narrative and documentation), see the report repo: https://github.com/Kenai-Watershed-Forum/kenai-river-wqx

------------------------------------------------------------------------

## Year Configuration

```{r year-config, message=FALSE, warning=FALSE}
# ── YEAR CONFIGURATION ─────────────────────────────────────────────────────────
# This is the ONLY block that must be updated for each new data year.
# All downstream scripts read from the `cfg` list constructed below.

library(tidyverse)
library(readxl)
library(openxlsx)
library(data.table)
library(stringr)
library(magrittr)
library(janitor)
library(hms)
library(lubridate)
library(anytime)
library(leaflet)

# ── Scalar date and year variables ─────────────────────────────────────────────
year <- XXXX  # e.g., 2023

# Field sampling dates for SGS/ALS/TSS ingestion
spring_sample_date <- "M/D/YYYY"  # e.g., "5/11/2021"
summer_sample_date <- "M/D/YYYY"  # e.g., "7/27/2021"

# Lab receipt dates (typically same day as collection)
spring_rec_date <- "YYYY-MM-DD"  # e.g., "2021-05-11"
summer_rec_date <- "YYYY-MM-DD"  # e.g., "2021-07-27"

# FC lab analysis dates (typically the day after field collection)
spring_fc_analysis_date <- mdy("M/D/YYYY")  # e.g., mdy("5/12/2021")
summer_fc_analysis_date <- mdy("M/D/YYYY")  # e.g., mdy("7/28/2021")

# FC lab source per season. Options: "SWWTP", "Taurianen"
# Note: Taurianen was used for summer FC in 2021 and some prior years.
#       Taurianen closed after 2021; 2022+ seasons use "SWWTP" for both.
spring_fc_lab <- "SWWTP"
summer_fc_lab <- "SWWTP"  # change to "Taurianen" for 2021 and applicable prior years

# Explicit FC file paths (not derived -- filenames embed sampling dates)
spring_fc_path <- "other/input/{year}_wqx_data/spring_{year}_wqx_data/SWWTP/[filename].xls"
summer_fc_path <- "other/input/{year}_wqx_data/summer_{year}_wqx_data/SWWTP/[filename].xlsx"

# Explicit TSS file paths (same reason as FC)
spring_tss_path <- "other/input/{year}_wqx_data/spring_{year}_wqx_data/SWWTP/[filename].xlsx"
summer_tss_path <- "other/input/{year}_wqx_data/summer_{year}_wqx_data/SWWTP/[filename].xlsx"

# ── Derived paths (update if folder naming conventions change) ─────────────────
data_dir        <- file.path("other/input", paste0(year, "_wqx_data"))
spring_data_dir <- file.path(data_dir, paste0("spring_", year, "_wqx_data"))
summer_data_dir <- file.path(data_dir, paste0("summer_", year, "_wqx_data"))
output_qaqc_dir <- file.path("other/output/lab_qaqc_data",
                              paste0(year, "_lab_qaqc_data"))

# Intermediate output paths
wqx_intermediate_path <- file.path("other/output/wqx_formatted/intermediate",
                                    paste0(year, "_kwf_baseline_results_wqx.csv"))
flagged_export_path   <- file.path("other/output/wqx_formatted/intermediate",
                                    paste0(year, "_export_data_flagged.csv"))
flag_decisions_path   <- file.path("other/input/wqx_templates",
                                    paste0(year, "_data_flag_decisions.csv"))

# ── cfg list (passed to stable function scripts) ───────────────────────────────
cfg <- list(
  year                  = year,
  templates_dir         = "other/input/wqx_templates",
  wqx_template_file     = file.path("other/input/wqx_templates",
                                     paste0("AWQMS_KWF_Baseline_", year, ".xlsx")),
  spring_data_dir       = spring_data_dir,
  summer_data_dir       = summer_data_dir,
  output_qaqc_dir       = output_qaqc_dir,
  wqx_intermediate_path = wqx_intermediate_path,
  flagged_export_path   = flagged_export_path,
  flag_decisions_path   = flag_decisions_path,
  wqx_downloads_dir     = "other/input/WQX_downloads",
  spring_sample_date    = spring_sample_date,
  summer_sample_date    = summer_sample_date
)

# Create output directories if they do not exist
dir.create(output_qaqc_dir,                             recursive = TRUE, showWarnings = FALSE)
dir.create(dirname(wqx_intermediate_path),              recursive = TRUE, showWarnings = FALSE)
dir.create("other/output/field_qa_qc_data/metals_total_diss", recursive = TRUE, showWarnings = FALSE)
dir.create("other/output/field_qa_qc_data/completeness_measures", recursive = TRUE, showWarnings = FALSE)
dir.create(file.path("other/output/lab_qaqc_data",
                     paste0(year, "_lab_qaqc_data")),   recursive = TRUE, showWarnings = FALSE)
dir.create("other/output/misc",                         recursive = TRUE, showWarnings = FALSE)
```

------------------------------------------------------------------------

## Part A: Data Ingestion

**ADAPT THIS SECTION FOR EACH NEW YEAR.**

Part A contains all lab-specific ingestion code. Because EDD formats vary by year and lab, this code is inlined here rather than sourced, so that year-specific adaptations are visible and explicitly marked. Use the scripts in `reference/` as worked examples from 2021.

After running all subsections, the final `dat` object must conform to the column contract documented in AGENTS.md before `source("functions/format_wqx.R")` will run correctly.

### SGS/ALS Lab Results

```{r ingest-sgs-als, message=FALSE, warning=FALSE, include=FALSE}

# ADAPT: Read spring and summer SGS EDD files.
# File naming convention follows: {season}_{year}_sgs_batch_info.csv
# See reference/2021_ingest_sgs_als.R for the complete 2021 approach.

# ADAPT: Check column names before renaming. SGS EDD column names may vary year to year.
# Run `names(spring_sgs_raw)` after read-in to verify.

# Read spring SGS EDD
spring_sgs_raw <- read.csv(
  file.path(cfg$spring_data_dir, "SGS", paste0("spring_", year, "_sgs_batch_info.csv"))
)

# Read summer SGS EDD
summer_sgs_raw <- read.csv(
  file.path(cfg$summer_data_dir, "SGS", paste0("summer_", year, "_sgs_batch_info.csv"))
)

# ADAPT: Bind, clean, rename columns to match the dat contract.
# Required output columns: sample, lab_sample, lab_name, analyte, result, units,
#   resultflag, loq, lod, collect_date, collect_time, rec_date, rec_time,
#   run_date, run_time, analytical_method, epa_analysis_id, context_code,
#   monitoring_location_id, note, sample_type, sample_condition, dissolved, matrix
#
# See reference/2021_ingest_sgs_als.R Parts A-E for the full transformation.

sgs_raw <- bind_rows(spring_sgs_raw, summer_sgs_raw) %>%
  clean_names() %>%
  remove_empty()

# ADAPT: Apply any year-specific unit corrections here.
# In 2021, dissolved Ca/Fe/Mg from SGS were labeled "ug/L" but were on the mg/L scale.
# Check the current year's EDD and PDF report for unit consistency before proceeding.

# ADAPT: Read ALS data (if applicable for this year).
# In 2021 ALS supplied dissolved Ca, Fe, Mg only. Check whether ALS was used this year.
# See reference/2021_ingest_sgs_als.R Part B.

# ADAPT: Join site-name matching table to resolve lab site name to WQX site IDs.
sgs_site_names <- read_excel(
  file.path(cfg$templates_dir, "sgs_site_names_matching_table_manual_edit.xlsx")
)

# ADAPT: Join analytical method matching table.
# Reference: other/input/wqx_templates/wqx_qaqc/ for method lookup tables.

# ADAPT: Segregate lab QA/QC rows (MB, LCS, DUP) from field results (PS).
# Lab QC goes to output_qaqc_dir; field results remain in dat.
# See reference/2021_ingest_sgs_als.R Parts D-E.

# At this point dat should contain only PS (project sample) rows.
# Assign sgs_dat <- dat before FC/TSS bind:
sgs_dat <- data.frame()  # REPLACE WITH ACTUAL INGESTION RESULT
```

### Fecal Coliform

```{r ingest-fc, message=FALSE, warning=FALSE, include=FALSE}

# ADAPT: FC ingestion depends on which lab was used (SWWTP or Taurianen).
# See reference/2021_ingest_fc.R for the complete 2021 approach with both parsers.

# ADAPT: SWWTP FC format differs from Taurianen format. The lab selector variables
# spring_fc_lab and summer_fc_lab (set in Year Configuration) determine which
# parser runs. For 2022+, both use SWWTP.

if (spring_fc_lab == "SWWTP") {
  # ADAPT: Read SWWTP spring FC file. Sheet name, skip rows, and column layout
  # may vary slightly year to year. Verify with names() after read.
  spring_fc_raw <- read_excel(spring_fc_path, skip = 2)
  # ... ADAPT: clean, rename, assign collect_date, analyte = "Fecal Coliform", etc.
  spring_fc <- data.frame()  # REPLACE WITH ACTUAL INGESTION RESULT
} else if (spring_fc_lab == "Taurianen") {
  # ADAPT: Taurianen format (2021 spring only -- see reference/2021_ingest_fc.R)
  spring_fc <- data.frame()  # REPLACE WITH ACTUAL INGESTION RESULT
}

if (summer_fc_lab == "SWWTP") {
  summer_fc_raw <- read_excel(summer_fc_path, skip = 2)
  # ... ADAPT
  summer_fc <- data.frame()  # REPLACE WITH ACTUAL INGESTION RESULT
} else if (summer_fc_lab == "Taurianen") {
  summer_fc <- data.frame()  # REPLACE WITH ACTUAL INGESTION RESULT
}

# Bind spring and summer FC and add to sgs_dat
fc_dat <- bind_rows(spring_fc, summer_fc)
```

### Total Suspended Solids

```{r ingest-tss, message=FALSE, warning=FALSE, include=FALSE}

# ADAPT: SWWTP TSS format (XLSX) is relatively consistent year to year, but
# verify sheet name, skip count, and column layout. See reference/2021_ingest_tss.R.

spring_tss_raw <- read_excel(spring_tss_path)
summer_tss_raw <- read_excel(summer_tss_path)

# ADAPT: Clean, assign collect_date (= rec_date for TSS), analyte = "Total suspended solids",
# units = "mg/L", analytical_method = "SM 2540 D", etc.
# Note: SWWTP did not report required lab QA results in 2021 and 2022 -- check for current year.

tss_dat <- data.frame()  # REPLACE WITH ACTUAL INGESTION RESULT
```

### Bind and Standardize

```{r bind-standardize, message=FALSE, warning=FALSE, include=FALSE}

# Bind all lab results into a single dataframe.
dat <- bind_rows(sgs_dat, fc_dat, tss_dat)

# ── dat column contract check ──────────────────────────────────────────────────
# Verify that all required columns are present before handing off to format_wqx.R.
# Missing columns will cause silent failures or errors downstream.
required_cols <- c(
  "monitoring_location_id", "collect_date", "collect_time",
  "analyte", "result", "units", "resultflag", "loq", "lod",
  "analytical_method", "epa_analysis_id", "context_code",
  "lab_name", "lab_sample", "run_date", "run_time",
  "note", "sample_type", "sample_condition"
)

missing_cols <- setdiff(required_cols, names(dat))
if (length(missing_cols) > 0) {
  stop("dat is missing required columns: ", paste(missing_cols, collapse = ", "))
}

cat("dat rows:", nrow(dat), "\n")
cat("Columns present:", paste(names(dat), collapse = ", "), "\n")
```

------------------------------------------------------------------------

## Part B: WQX Formatting

```{r format-wqx, message=FALSE, warning=FALSE, include=FALSE}

# Applies all lookup table joins (coordinates, sample fractions, detection condition,
# preservative, container type), constructs WQX column names, builds Activity IDs,
# and writes the provisional WQX intermediate CSV.
#
# Requires: dat, cfg (set in Year Configuration)
# Produces: dat (raw cols restored), dat_raw, all_dat; writes cfg$wqx_intermediate_path

source("functions/format_wqx.R")
```

### Provisional Results (Prior to QA/QC Review)

The following table summarizes results after WQX formatting, before QA/QC evaluation.

```{r provisional-summary, message=FALSE}
read.csv(cfg$wqx_intermediate_path) %>%
  clean_names() %>%
  count(characteristic_name, result_analytical_method_id, activity_start_date) %>%
  arrange(activity_start_date, characteristic_name)
```

------------------------------------------------------------------------

## Part C: QA/QC Checklist

Prior to upload to the EPA WQX, all water quality data is checked against a standard Data Evaluation Checklist developed in coordination with the Alaska Department of Environmental Conservation.

#### Pre-Database

##### Overall Project Success

**1.) Were the appropriate analytical methods used for all parameters?**

```{r q01, message=FALSE}
# MANUAL: Confirm that analytical methods match those in the approved QAPP.
# List any deviations and their disposition.

# "Yes. Analytical methods from the approved [year] QAPP were employed. [any exceptions]"
```

<br>

**2.) Were there any deviations from the sampling plan?**

```{r q02}
# MANUAL: Were all sites visited? Were all planned sampling dates completed?
# Note any instrument or method deviations.
```

<br>

**3.) Were field duplicates, blanks, and/or other QC samples collected as planned?**

```{r q03, message=FALSE, warning=FALSE}
# Compare planned vs. actual samples. Reads from the SGS bottle order spreadsheet
# (KRBWQM_SGS_bottle_order_{season}_{year}.xlsx) and the WQX intermediate CSV.

# ADAPT: Sheet names within the bottle order workbooks may differ by year.
# ADAPT: The analysis name -> method ID mapping (case_when below) may need updating
# if method IDs or analysis names changed in the current year.

total_samples_collected_summary <- read.csv(cfg$wqx_intermediate_path) %>%
  clean_names() %>%
  group_by(result_analytical_method_id, activity_start_date, activity_type) %>%
  count() %>%
  rename(actual_results_n = n) %>%
  transform(activity_start_date = ymd(activity_start_date))

spring_planned <- read_excel(
  file.path(cfg$spring_data_dir, "SGS",
            paste0("KRBWQM_SGS_bottle_order_Spring_", year, ".xlsx")),
  sheet = paste0("bottles_analyses_spring_", year)
) %>% mutate(activity_start_date = spring_sample_date)

summer_planned <- read_excel(
  file.path(cfg$summer_data_dir, "SGS",
            paste0("KRBWQM_SGS_bottle_order_Summer_", year, ".xlsx")),
  sheet = paste0("bottles_analyses_summer_", year)
) %>% mutate(activity_start_date = summer_sample_date)

planned_samples <- bind_rows(spring_planned, summer_planned) %>%
  clean_names() %>%
  filter(item == "Bottle") %>%
  select(monitoring_location_id, site, container_size, quantity,
         preservative, analysis_1, analysis_2, activity_start_date) %>%
  fill(site, .direction = "down") %>%
  fill(monitoring_location_id, .direction = "down") %>%
  filter(site != "Spare Kit") %>%
  mutate(sample_type = case_when(
    grepl("DUPLICATE", site, ignore.case = TRUE) ~ "FIELD DUPLICATE",
    grepl("blank", analysis_1, ignore.case = TRUE) ~ "TRIP BLANK"
  )) %>%
  pivot_longer(values_to = "analysis", cols = c("analysis_1", "analysis_2")) %>%
  filter(!is.na(analysis)) %>%
  mutate(analysis = str_replace(analysis, "624 - BTEX trip blank", "624 - BTEX")) %>%
  # ADAPT: Confirm analysis name -> method ID mappings for the current year
  mutate(analysis = case_when(
    analysis == "SM4500-NO3E - Nitrogen (Nitrate+Nitrite)" ~ "4500-NO3(F)",
    analysis == "SM4500-PE - Total Phosphorus"             ~ "4500-P-E",
    analysis == "200.7 - Total Metals"                    ~ "200.7",
    analysis == "200.8 - Dissolved Metals"                 ~ "200.8",
    analysis == "1 liter Total Suspended Solids"           ~ "2540-D",
    analysis == "Fecal Coliform"                           ~ "9222D",
    analysis == "624 - BTEX"                               ~ "8260D",
    TRUE ~ analysis
  )) %>%
  # ADAPT: Update expected_results counts if the analyte lists changed
  mutate(expected_results = case_when(
    analysis == "4500-NO3(F)" ~ 1,
    analysis == "4500-P-E"   ~ 1,
    analysis == "200.7"      ~ 3,
    analysis == "200.8"      ~ 6,
    analysis == "2540-D"     ~ 1,
    analysis == "9222D"      ~ 1,
    analysis == "8260D"      ~ 6
  )) %>%
  mutate(activity_type = case_when(
    sample_type == "FIELD DUPLICATE" ~ "Field Duplicate",
    sample_type == "TRIP BLANK"      ~ "Trip Blank",
    TRUE ~ "Field Msr/Obs"
  )) %>%
  select(-sample_type)

planned_samples_summary <- planned_samples %>%
  group_by(analysis, expected_results, activity_start_date, activity_type) %>%
  summarise(expected_ct = sum(expected_results), .groups = "drop") %>%
  rename(expected_results_n = expected_ct,
         result_analytical_method_id = analysis) %>%
  transform(activity_start_date = mdy(activity_start_date),
            expected_results_n  = as.numeric(expected_results_n)) %>%
  mutate(activity_type = case_when(
    activity_type == "Field Duplicate" ~ "Quality Control Field Replicate Msr/Obs",
    activity_type == "Field Msr/Obs"   ~ "Field Msr/Obs",
    activity_type == "Trip Blank"      ~ "Quality Control Sample-Trip Blank"
  ))

planned_samples_summary_by_site <- planned_samples %>%
  group_by(monitoring_location_id, analysis, expected_results, activity_start_date, activity_type) %>%
  summarise(expected_ct = sum(expected_results), .groups = "drop") %>%
  rename(expected_results_n = expected_ct,
         result_analytical_method_id = analysis) %>%
  transform(activity_start_date = mdy(activity_start_date),
            expected_results_n  = as.numeric(expected_results_n)) %>%
  mutate(activity_type = case_when(
    activity_type == "Field Duplicate" ~ "Quality Control Field Replicate Msr/Obs",
    activity_type == "Field Msr/Obs"   ~ "Field Msr/Obs",
    activity_type == "Trip Blank"      ~ "Quality Control Sample-Trip Blank"
  ))

planned_actual_analyses <- left_join(
  planned_samples_summary, total_samples_collected_summary,
  by = c("result_analytical_method_id", "activity_start_date", "activity_type")
) %>%
  mutate(pct_diff = ((actual_results_n - expected_results_n) /
                       (actual_results_n + expected_results_n)) * 100)

write.csv(planned_actual_analyses,
          file.path("other/output/field_qa_qc_data",
                    paste0("planned_actual_analyses_", year, ".csv")),
          row.names = FALSE)

planned_actual_analyses
```

```{r q03-narrative}
# MANUAL: Describe any deviations between planned and actual results.
# Identify root causes (chain of custody errors, lab errors, missed sites, etc.)
```

<br>

**4.) Do the laboratory reports provide results for all sites and parameters?**

```{r q04}
# MANUAL: Confirm based on Q3 results. Note any exceptions.
```

<br>

**5.) Is a copy of the Chain of Custody included with the laboratory reports?**

```{r q05}
# MANUAL: Describe CoC status for each contracted lab this year.
```

<br>

**6.) Do the laboratory reports match the Chain of Custody and requested methods throughout?**

```{r q06}
# MANUAL: Note any mismatches. Reference Q3 for method-level deviations.
```

<br>

**7.) Are the number of samples on the laboratory reports the same as on the Chain of Custody?**

```{r q07}
# MANUAL: Confirm sample counts match between CoC and lab report.
```

<br>

**8.) Was all supporting info provided in the laboratory report, such as reporting limits for all analyses and definitions?**

```{r q08}
# MANUAL: Note reporting limit availability per lab. SWWTP limits come from the QAPP.
```

<br>

**9.) Are site names, dates, and times correct and as expected?**

```{r q09, message=FALSE}
# Formulaic check: review unique site IDs in WQX intermediate against known site list.
sites_in_data <- read.csv(cfg$wqx_intermediate_path) %>%
  clean_names() %>%
  distinct(monitoring_location_id, activity_start_date) %>%
  arrange(monitoring_location_id, activity_start_date)

sites_in_data

# MANUAL: Confirm all expected sites are present. Note any transcription errors
# corrected during ingestion (document the correction method in Part A comments).
```

<br>

**10.) Were there any issues with instrument calibration?**

```{r q10}
# MANUAL: Describe calibration status for all field instruments used.
# For sonde-based field parameters: note calibration records.
# For hand-held probes: note verification method (e.g., ice bath for temperature).
```

<br>

**11.) Did the instruments perform as expected?**

```{r q11}
# MANUAL: Note any instrument failures or anomalous readings.
```

<br>

**12.) Was instrument calibration performed according to the QAPP and instrument recommendations?**

```{r q12}
# MANUAL: Confirm or note exceptions.
```

<br>

**13.) Was instrument verification during the field season performed according to the QAPP and instrument recommendations?**

```{r q13}
# MANUAL: Confirm or note exceptions.
```

<br>

**14.) Were instrument calibration verification logs or records kept?**

```{r q14}
# MANUAL: Confirm records are on file. Note location.
```

<br>

**15.) Do the instrument data files site IDs, time stamps and file names match?**

```{r q15}
# MANUAL: If instrument files were used, confirm. If not, note.
```

<br>

**16.) Is any insitu field data rejected and why?**

```{r q16}
# MANUAL: List any rejected in-situ measurements and rationale.
```

<br>

**17.) Were preservation, hold time and temperature requirements met?**

```{r q17-holdtimes, message=FALSE, warning=FALSE}
# Formulaic: calculate hold time from collection to receipt and compare to max allowed.
col_names <- c("sample", "epa_analysis_id", "analyte", "collect_date", "collect_time",
               "rec_date", "rec_time", "lab_name")

holdtime_dat <- dat %>%
  select(any_of(col_names)) %>%
  mutate(rec_date1 = as.Date(rec_date),
         rec_time1 = case_when(
           !is.na(rec_time) ~ as_hms(rec_time),
           is.na(rec_time)  ~ as_hms(rec_date)
         )) %>%
  select(-rec_time, -rec_date) %>%
  rename(rec_time = rec_time1, rec_date = rec_date1) %>%
  mutate(activity_datetime = ymd_hms(paste(collect_date, collect_time)),
         rec_datetime       = ymd_hms(paste(rec_date, rec_time)),
         hold_time_hours    = as.numeric(rec_datetime - activity_datetime)) %>%
  select(-collect_date, -collect_time, -rec_date, -rec_time) %>%
  mutate(analytical_method = str_trim(dat$analytical_method, side = "left"))

max_holding_times <- read.csv(
  file.path(cfg$templates_dir, "sample_holding_times.csv")
) %>% transform(max_holding_time_hours = as.numeric(max_holding_time_hours))

holdtime_dat <- left_join(holdtime_dat, max_holding_times) %>%
  mutate(hold_time_pass = case_when(
    hold_time_hours > max_holding_time_hours ~ "N",
    TRUE ~ "Y"
  ))

write.csv(holdtime_dat,
          "other/output/field_qa_qc_data/holding_time_calcs.csv",
          row.names = FALSE)

cat("Hold time failures:", sum(holdtime_dat$hold_time_pass == "N", na.rm = TRUE), "\n")
holdtime_dat %>% filter(hold_time_pass == "N")
```

<br>

**18.) Are dissolved metal quantities less than total metals quantities?**

```{r q18-dissolved-total, message=FALSE, warning=FALSE}
# Formulaic: compare dissolved vs. total for Ca, Fe, Mg (where both are present).
# ADAPT: Adjust element list if different parameters are available this year.

diss_total_analytes <- c("Calcium, Total", "Iron, Total", "Magnesium, Total",
                         "Calcium", "Iron", "Magnesium")

diss_total_metals_dat <- dat %>%
  select(analyte, collect_date, collect_time, monitoring_location_id,
         result, units, result_detection_condition, sample_condition) %>%
  filter(analyte %in% diss_total_analytes) %>%
  mutate(result = case_when(
    units == "mg/l" ~ result,
    units == "ug/L" ~ result / 1000,
    TRUE ~ result
  ), units = "mg/L")

dir_diss <- "other/output/field_qa_qc_data/metals_total_diss/"

for (elem in c("Calcium", "Iron", "Magnesium")) {
  elem_dat <- diss_total_metals_dat %>%
    filter(grepl(elem, analyte)) %>%
    pivot_wider(names_from = "analyte", values_from = "result") %>%
    mutate(particulate_conc = .data[[paste0(elem, ", Total")]] - .data[[elem]],
           total_greater_diss = ifelse(particulate_conc > 0, "Y", "N"))
  write.csv(elem_dat, file.path(dir_diss, paste0(tolower(elem), ".csv")))
  cat(elem, "rows with total > dissolved:",
      sum(elem_dat$total_greater_diss == "Y", na.rm = TRUE),
      "of", nrow(elem_dat), "\n")
}
```

```{r q18-narrative}
# MANUAL: Interpret the dissolved vs. total comparison results.
# Note any systematic offsets (as seen in 2021 Ca) and their potential causes.
```

<br>

**19.) Are the duplicate sample(s) RPD (Relative Percent Difference) within range described in QAPP?**

```{r q19-rpd, message=FALSE, warning=FALSE}
# Formulaic: calculate RPD for field duplicate pairs at QC sites.
# ADAPT: Update the qaqc info sheet for the current year's QC sites.

qaqc_sites <- read_excel(
  file.path(cfg$templates_dir, "wqx_qaqc/wqx_qaqc_info.xlsx"),
  sheet = paste0("field_dup_sites_", year)
) %>% remove_empty()

export_dat_rpd <- read.csv(cfg$wqx_intermediate_path) %>% clean_names()
field_dup_dat  <- inner_join(export_dat_rpd, qaqc_sites)

rpd_cols <- c("monitoring_location_id", "activity_type", "activity_start_date",
              "characteristic_name", "result_detection_condition",
              "result_value", "result_unit",
              "result_detection_limit_type_1", "result_detection_limit_value_1",
              "result_detection_limit_unit_1")

rpd_check_dat <- field_dup_dat %>%
  select(all_of(rpd_cols)) %>%
  pivot_wider(names_from = "activity_type", values_from = "result_value") %>%
  clean_names() %>%
  mutate(
    field_above_loq = field_msr_obs > result_detection_limit_value_1,
    field_loq_2x    = field_msr_obs > 2 * result_detection_limit_value_1,
    qc_above_loq    = quality_control_field_replicate_msr_obs > result_detection_limit_value_1,
    qc_loq_2x       = quality_control_field_replicate_msr_obs > 2 * result_detection_limit_value_1
  ) %>%
  mutate(
    loq_pass    = field_above_loq | qc_above_loq,
    loq_2x_pass = field_loq_2x | qc_loq_2x,
    rpd_eligible = loq_pass & loq_2x_pass &
      !is.na(field_msr_obs) & !is.na(quality_control_field_replicate_msr_obs)
  ) %>%
  mutate(
    rpd_pct = abs(
      ((field_msr_obs - quality_control_field_replicate_msr_obs) /
         ((field_msr_obs + quality_control_field_replicate_msr_obs) / 2)) * 100
    )
  )

# ADAPT: Update QAPP thresholds if they changed for the current year.
rpd_eligible_dat <- rpd_check_dat %>%
  filter(rpd_eligible) %>%
  mutate(threshold = case_when(
    grepl("Nitrate|Nitrite",  characteristic_name, ignore.case = TRUE) ~ 25,
    grepl("Phosphorus",       characteristic_name, ignore.case = TRUE) ~ 25,
    grepl("suspended|solids", characteristic_name, ignore.case = TRUE) ~ 5,
    TRUE ~ 20
  )) %>%
  filter(!is.na(rpd_pct))

rpd_exceed <- rpd_eligible_dat %>% filter(rpd_pct > threshold)

write.csv(rpd_check_dat, "other/output/field_qa_qc_data/rpd_check_dat.csv", row.names = FALSE)

cat("Total duplicate pairs:", nrow(rpd_check_dat), "\n")
cat("RPD-eligible pairs:", sum(rpd_check_dat$rpd_eligible, na.rm = TRUE), "\n")
cat("Pairs exceeding QAPP threshold:", nrow(rpd_exceed), "\n")
```

```{r q19-narrative}
# MANUAL: Discuss RPD results per parameter. Follow the ADEC 2021 approach:
# below-LOQ pairs are retained as Accepted non-detects, not flagged.
# Identify any parameters that will be flagged based on RPD failures.
```

<br>

**20.) Were there any laboratory discrepancies, errors, data qualifiers, or QC failures?**

```{r q20-matrix-spikes, message=FALSE, warning=FALSE}
# Formulaic: review matrix spike recovery from SGS lab QA/QC data.
matrix_spike_limits <- read.csv(
  file.path(cfg$output_qaqc_dir, "sgs_als_qaqc_dat.csv")
) %>%
  select(lab_name, sample, collect_date, sample_type, result, analyte,
         analytical_method, resultflag, percent_recovered,
         rec_limit_low, rec_limit_high, sample_rpd, rpd_limit_low, rpd_limit_high,
         loq, lod, detection_limit, sample_condition) %>%
  mutate(rec_limit_pass = case_when(
    is.na(rec_limit_low) ~ "",
    percent_recovered > rec_limit_high | percent_recovered < rec_limit_low ~ "N",
    TRUE ~ "Y"
  ))

matrix_spike_fails <- matrix_spike_limits %>% filter(rec_limit_pass == "N")

write.csv(matrix_spike_fails,
          file.path("other/output/lab_qaqc_data",
                    paste0("matrix_spike_recovery_fails_", year, ".csv")),
          row.names = FALSE)

cat("Matrix spike recovery failures:", nrow(matrix_spike_fails), "\n")
if (nrow(matrix_spike_fails) > 0) matrix_spike_fails
```

```{r q20-narrative}
# MANUAL: Describe any additional lab QC anomalies not captured by the matrix spike check.
# Review the lab PDF reports for narrative QC notes.
```

<br>

**21.) Is any laboratory data rejected and why?**

```{r q21}
# MANUAL: List parameters to be flagged based on Q19-Q20 findings.
# Flagging decisions are recorded in cfg$flag_decisions_path ({year}_data_flag_decisions.csv).
# That CSV is joined in Part D (apply_qaqc_flags.R).
```

<br>

**22.) During the field season, review raw data files (EDDs, instrument records) as they are received. Document changes and corrections to methods as needed.**

```{r q22}
# MANUAL: Describe the EDD review process for the current year.
# Note any corrections applied during ingestion and how they were documented.
```

<br>

**23.) Is the dataset complete and did you receive the expected number of results?**

```{r q23}
# MANUAL: Reference Q3. Summarize deviations and their known causes.
```

<br>

**24.) Was the data collected representative of environmental conditions?**

```{r q24}
# MANUAL: Describe representativeness. Note any flagged data that may not be representative.
```

<br>

**25.) Does project meet Completeness Measure A criteria?**

From the QAPP, CMA = primary samples collected / usable samples submitted (goal: 85%).

```{r apply-flags, message=FALSE, warning=FALSE}
# Join flag decisions and write flagged export CSV.
# IMPORTANT: This chunk must run before Q25 and Q26 calculations.
# Requires: cfg$wqx_intermediate_path, cfg$flag_decisions_path, cfg$flagged_export_path
source("functions/apply_qaqc_flags.R")
```

```{r q25-cma, message=FALSE}
analytes_qapp <- read.csv(
  file.path(cfg$templates_dir, "analytes_list_manual_edit.csv")
)
colnames(analytes_qapp) <- c("characteristic_name", "a", "param_in_qapp")
analytes_qapp <- analytes_qapp %>% select(-a)

# BY PARAMETER
completeness_summary_param <- read.csv(cfg$flagged_export_path) %>%
  clean_names() %>%
  left_join(analytes_qapp) %>%
  filter(param_in_qapp == "Y") %>%
  group_by(result_analytical_method_id, characteristic_name, flag) %>%
  count() %>%
  rename(actual_results_n = n) %>%
  pivot_wider(names_from = "flag", values_from = "actual_results_n") %>%
  rename(flag_Y = Y, flag_N = N) %>%
  rowwise() %>%
  mutate(total_samples = sum(flag_Y, flag_N, na.rm = TRUE),
         cma = flag_N / total_samples,
         cma = replace_na(cma, 0)) %>%
  adorn_pct_formatting(,,,cma)
write.csv(completeness_summary_param,
          "other/output/field_qa_qc_data/completeness_measures/cma_parameter.csv")

# BY SITE
completeness_summary_site <- read.csv(cfg$flagged_export_path) %>%
  clean_names() %>%
  left_join(analytes_qapp) %>%
  filter(param_in_qapp == "Y") %>%
  group_by(monitoring_location_id, flag) %>%
  count() %>%
  rename(actual_results_n = n) %>%
  pivot_wider(names_from = "flag", values_from = "actual_results_n") %>%
  rename(flag_Y = Y, flag_N = N) %>%
  rowwise() %>%
  mutate(total_samples = sum(flag_Y, flag_N, na.rm = TRUE),
         cma = flag_N / total_samples,
         cma = replace_na(cma, 0)) %>%
  adorn_pct_formatting(,,,cma)
write.csv(completeness_summary_site,
          "other/output/field_qa_qc_data/completeness_measures/cma_site.csv")

# OVERALL
completeness_summary_project <- read.csv(cfg$flagged_export_path) %>%
  clean_names() %>%
  left_join(analytes_qapp) %>%
  filter(param_in_qapp == "Y") %>%
  group_by(flag) %>%
  count() %>%
  pivot_wider(names_from = "flag", values_from = "n") %>%
  rename(flag_Y = Y, flag_N = N) %>%
  rowwise() %>%
  mutate(total_samples = sum(flag_Y, flag_N, na.rm = TRUE),
         cma = flag_N / total_samples,
         cma = replace_na(cma, 0)) %>%
  adorn_pct_formatting(,,,cma)
write.csv(completeness_summary_project,
          "other/output/field_qa_qc_data/completeness_measures/cma_project.csv")

# Summary scalars
cma_dat_inline   <- read.csv(cfg$flagged_export_path) |> clean_names() |>
  left_join(analytes_qapp) |> filter(param_in_qapp == "Y")
cma_total        <- nrow(cma_dat_inline)
cma_unflagged_n  <- sum(cma_dat_inline$flag == "N", na.rm = TRUE)
cma_flagged_n    <- sum(cma_dat_inline$flag == "Y", na.rm = TRUE)
cma_overall_pct  <- paste0(round(cma_unflagged_n / cma_total * 100, 1), "%")

cat("CMA (overall):", cma_overall_pct, "(goal: 85%)\n")
completeness_summary_param
```

<br>

**26.) Does project meet Completeness Measure B criteria?**

From the QAPP, CMB = unflagged results / planned results (goal: 60%).

```{r q26-cmb, message=FALSE}
# BY PARAMETER
cmb_by_method <- read.csv(cfg$flagged_export_path) %>%
  clean_names() %>%
  left_join(analytes_qapp) %>%
  filter(param_in_qapp == "Y") %>%
  group_by(result_analytical_method_id, activity_start_date, activity_type, flag) %>%
  count() %>%
  rename(unflagged_results_n = n) %>%
  pivot_wider(names_from = "flag", values_from = "unflagged_results_n") %>%
  rename(flag_Y = Y, flag_N = N) %>%
  transform(activity_start_date = ymd(activity_start_date)) %>%
  left_join(planned_samples_summary) %>%
  mutate(cmb = flag_N / expected_results_n,
         cmb = replace_na(cmb, 0)) %>%
  adorn_pct_formatting(,,,cmb)
write.csv(cmb_by_method, "other/output/field_qa_qc_data/completeness_measures/cmb_parameter.csv")

# BY SITE
planned_by_site_totals <- planned_samples_summary_by_site %>%
  group_by(monitoring_location_id) %>%
  summarise(expected_results_n = sum(expected_results_n), .groups = "drop")

cmb_by_site <- read.csv(cfg$flagged_export_path) %>%
  clean_names() %>%
  left_join(analytes_qapp) %>%
  filter(param_in_qapp == "Y") %>%
  group_by(monitoring_location_id, flag) %>%
  count() %>%
  rename(unflagged_results_n = n) %>%
  pivot_wider(names_from = "flag", values_from = "unflagged_results_n") %>%
  rename(flag_Y = Y, flag_N = N) %>%
  left_join(planned_by_site_totals) %>%
  mutate(cmb = flag_N / expected_results_n,
         cmb = replace_na(cmb, 0)) %>%
  adorn_pct_formatting(,,,cmb)
write.csv(cmb_by_site, "other/output/field_qa_qc_data/completeness_measures/cmb_site.csv")

# BY PROJECT WHOLE
total_planned <- sum(planned_samples_summary$expected_results_n)
cmb_project <- read.csv(cfg$flagged_export_path) %>%
  clean_names() %>%
  left_join(analytes_qapp) %>%
  filter(param_in_qapp == "Y") %>%
  group_by(flag) %>%
  count() %>%
  pivot_wider(names_from = "flag", values_from = "n") %>%
  rename(flag_Y = Y, flag_N = N) %>%
  mutate(total_planned = total_planned,
         cmb = flag_N / total_planned,
         cmb = replace_na(cmb, 0)) %>%
  adorn_pct_formatting(,,,cmb)
write.csv(cmb_project, "other/output/field_qa_qc_data/completeness_measures/cmb_project.csv")

cmb_usable  <- cmb_project$flag_N
cmb_overall_pct <- paste0(round(cmb_usable / total_planned * 100, 1), "%")
cat("CMB (overall):", cmb_overall_pct, "(goal: 60%)\n")
cmb_by_method
```

<br>

**27.) Was the QA officer consulted for any data concerns?**

```{r q27}
# MANUAL: Describe any consultations with KWF QA staff or ADEC.
```

<br>

**28.) Are the correct monitoring locations associated with the project? Are latitude and longitude filled out in a consistent format?**

```{r q28-map, message=FALSE, warning=FALSE}
export_dat <- read.csv(cfg$flagged_export_path)

site_check <- export_dat %>%
  distinct(monitoring_location_id, activity_latitude, activity_longitude) %>%
  transform(activity_longitude = as.numeric(
    iconv(activity_longitude, "utf-8", "ascii", sub = "")
  )) %>%
  # ADAPT: Check for any sites with sign-flipped longitude (a known issue for KR RM 43)
  mutate(activity_longitude = case_when(
    activity_longitude > 0 ~ activity_longitude * -1,
    TRUE ~ activity_longitude
  ))

if (knitr::is_html_output()) {
  leaflet(site_check) %>%
    addTiles() %>%
    addMarkers(~activity_longitude, ~activity_latitude,
               popup = ~as.character(monitoring_location_id),
               label  = ~as.character(monitoring_location_id))
} else {
  knitr::kable(site_check, caption = paste(year, "monitoring locations"))
}
```

<br>

**29.) Are the QAPP and other supporting documents attached?**

```{r q29}
# MANUAL: Confirm that the current approved QAPP will be included in the CDX submission.
# The QAPP filename is set in generate_cdx_export.R (ProjectAttachmentFileName).
```

<br>

**30.) Is all project metadata correct?**

```{r q30}
# MANUAL: Confirm organization ID, project ID, QAPP approval status, site coordinates.
```

<br>

**31.) Is the organization ID correct?**

```{r q31}
# Organization ID for KWF: "Kenai_WQX" -- confirmed manually during each CDX upload.
cat("Expected organization ID: Kenai_WQX\n")
```

<br>

**32.) Are the time zones consistent and correct (AKDT in summer)?**

```{r q32-timezone}
# Add time_zone column to export_dat and re-write flagged export.
# ADAPT: Confirm "AKDT" is correct for the current year's sampling dates.
# Spring sampling during AKST (before second Sunday of March) would require "AKST".
export_dat <- read.csv(cfg$flagged_export_path) %>%
  mutate(time_zone = "AKDT")
write.csv(export_dat, cfg$flagged_export_path)
cat("Time zone applied: AKDT\n")
```

<br>

**33.) Are all media types included? Media types appropriate to Characteristic?**

```{r q33}
cat("Activity media types in dataset:\n")
unique(export_dat$activity_media_name)
# "Water" is the only media type in this project.
```

<br>

**34.) Check Sample Collection, Preparation and Preservation Methods, Thermal Preservative, Equipment ID, Activity Media.**

```{r q34}
cat("Equipment IDs:\n")
unique(export_dat$equipment_id)
cat("\nResult sample fractions:\n")
unique(export_dat$result_sample_fraction)
```

<br>

**35.) Are all expected activity types present and are QC samples correctly identified?**

```{r q35}
cat("Activity types:\n")
unique(export_dat$activity_type)
```

<br>

**36.) Is the Activity media subdivision filled in (if relevant)?**

```{r q36}
cat("Activity media subdivision names:\n")
unique(export_dat$activity_media_subdivision_name)
```

<br>

**37.) For Water activity media, is the relative depth filled in?**

```{r q37}
cat("Depth/height measure values:\n")
unique(export_dat$activity_depth_height_measure)
# All surface grab samples recorded at ~15 cm depth.
```

<br>

**38.) Is the number of results for each Characteristic correct?**

```{r q38}
chr_num <- export_dat %>%
  group_by(characteristic_name) %>%
  tally()

dir_misc <- file.path("other/output/misc", paste0("characteristic_count_", year, ".csv"))
write.csv(chr_num, dir_misc, row.names = FALSE)
chr_num
```

<br>

**39.) Do the range of result values make sense?**

```{r q39}
# MANUAL: Evaluate result ranges visually and numerically.
# Compare against prior years and regulatory limits.
# Document likely outliers in other/input/outliers/baseline_outliers.xlsx.

# Quick range check by characteristic:
export_dat %>%
  filter(!is.na(result_value)) %>%
  group_by(characteristic_name, result_unit) %>%
  summarise(min = min(result_value, na.rm = TRUE),
            max = max(result_value, na.rm = TRUE),
            n   = n(), .groups = "drop") %>%
  arrange(characteristic_name)
```

<br>

**40.) Are units correct and consistent for each parameter?**

```{r q40}
# Check for mixed unit labels (e.g., "mg/L" vs. "mg/l").
z <- export_dat %>%
  select(characteristic_name, result_unit) %>%
  distinct()

cat("Unique unit values:\n")
print(unique(z$result_unit))

# Standardize case if needed
export_dat[export_dat == "mg/l"] <- "mg/L"
write.csv(export_dat, cfg$flagged_export_path)
```

<br>

**41.) Are detection limits and laboratory qualifiers included for analytical results?**

```{r q41}
cat("Result qualifiers present:\n")
unique(export_dat$result_qualifier)
cat("\nDetection limit types present:\n")
unique(export_dat$result_detection_limit_type_1)
```

<br>

**42.) Are results trip blanks and/or field blanks above detection limits, and is any related data flagged as a result?**

```{r q42}
# Identify trip blank results and check for detections.
trip_blanks <- export_dat %>%
  filter(activity_type == "Quality Control Sample-Trip Blank")

cat("Trip blank results:", nrow(trip_blanks), "\n")
cat("Detections in trip blanks (result_value > 0):",
    sum(!is.na(trip_blanks$result_value) & trip_blanks$result_value > 0), "\n")

# MANUAL: If any trip blank analytes are detected, determine whether associated
# field results should be flagged and update flag_decisions_path accordingly.
```

------------------------------------------------------------------------

## Part D: Flag + CDX Export

```{r generate-cdx, message=FALSE, warning=FALSE}

# Apply Result Status ID (Accepted/Rejected), finalize column format, and write
# three CDX upload files: results_activities.csv, project.csv, station.csv.
#
# REVIEW: generate_cdx_export.R contains a filter that excludes Phosphorus from
# EPA Method 200.8 (not in the 2021 QAPP). Confirm whether this filter applies
# to the current year before running. See the REVIEW comment in that script.
#
# Requires: export_dat (with time_zone applied in Q32 above), cfg

source("functions/generate_cdx_export.R")
```

### CDX Upload Files

The following files are ready for upload to the EPA Water Quality Exchange Central Data Exchange (CDX) at https://cdx.epa.gov/:

```{r cdx-file-summary, message=FALSE}
cdx_files <- c(
  "other/output/wqx_formatted/results_activities.csv",
  "other/output/wqx_formatted/project.csv",
  "other/output/wqx_formatted/station.csv"
)

for (f in cdx_files) {
  if (file.exists(f)) {
    d <- read.csv(f)
    cat(basename(f), ":", nrow(d), "rows\n")
  } else {
    cat(basename(f), ": FILE NOT FOUND\n")
  }
}
```

**Upload notes for `r year`:**

```{r cdx-notes}
# MANUAL: Document any year-specific notes relevant to the CDX upload:
# - Any parameters excluded and why
# - Upload date and CDX submission ID (once uploaded)
# - Any corrections made post-upload
# - Status of EPA ETL processing
```

\newpage
