---
execute:
  echo: false
  cache: false
date: "`r Sys.Date()`"
format:
  html:
    code-fold: true
    code-tools: true
    code-summary: "Show code"
---

# Appendix: Data Review and Upload {#appendix_a}

## Introduction

Prior to publishing analysis and interpretation of water quality data, we will ensure that all data that meets QA/QC standards outlined in the current project [Quality Assurance Project Plan (QAPP)](https://paperpile.com/app/p/7703451b-460d-00b4-82a0-1086ea2554c3) and is accessible in the appropriate public repository.

Water quality data from this project is ultimately destined for the Environmental Protection Agency's Water Quality Exchange (EPA WQX).

The Quality Assurance Project Plan (QAPP) for this project describes data management details and responsible parties for each step of the data pipeline from observation to repository. The 2021 data preparation and review process is published here as an appendix as an example of the process applied annually to each year's data.

Year 2021 QA/QC is included here as an example. Other project years QA/QC may be accessed at <https://kenai-watershed-forum.github.io/kenai-river-wqx-qaqc/>.

### Year 2021 Water Quality Data

In this appendix we will collate 2021 laboratory data from several sources into a single spreadsheet document with a consistent format. The desired end format is a spreadsheet template provided by the EPA Water Quality Exchange. These template files are available to download from the EPA at <https://www.epa.gov/waterdata/water-quality-exchange-web-template-files>.

Once the data is collated, it will be evaluated according to a Quality Assurance Checklist (template example provided by the Alaska Department of Environmental Conservation Soldotna office). Field observations that do not meet the quality assurance standards described in the evaluation checklist will be flagged before being uploaded to the EPA WQX.

Data that has been uploaded to the EPA WQX is evaluated biannually by the Alaska Department of Environmental Conservation (ADEC) in their [Integrated Water Quality Monitoring and Assessment Report](https://dec.alaska.gov/water/water-quality/integrated-report/)[^appendix_a-1]. The integrated report evaluates available water quality data from the previous five years against Alaska water quality standards and regulations [@adec2020].

[^appendix_a-1]: https://dec.alaska.gov/water/water-quality/integrated-report/

#### 2021 Water Quality Data WQX Formatting

The code scripts in this document assemble water quality data from the three analytical laboratories that partnered with Kenai Watershed Forum for this project in 2021:

-   SGS Laboratories (Anchorage, AK)

-   Soldotna Wastewater Treatment Plant (Soldotna, AK)

-   Taurianen Engineering and Testing (Soldotna, AK)

<br>

------------------------------------------------------------------------

##### 2021 Metals/Nutrients Lab Results (SGS Labs)

```{r, echo = F, message = F}
library("xfun")
source("functions/embed_if_html.R")

embed_if_html('other/input/2021_wqx_data/spring_2021_wqx_data/SGS/spring_2021_sgs_batch_info.csv', text = "Download Original Spring 2021 Metals/Nutrients Lab Results from SGS - Electronic Data Delivery file")

embed_if_html('other/input/2021_wqx_data/spring_2021_wqx_data/SGS/Spring 2021 Results from SGS.pdf', text = "Download Original Spring 2021 Metals/Nutrients Lab Results from SGS - PDF file")

```

```{r, echo = F}
embed_if_html('other/input/2021_wqx_data/summer_2021_wqx_data/SGS/summer_2021_sgs_batch_info.csv', text = "Download Original Summer 2021 Metals/Nutrients Lab Results from SGS - Electronic Data Delivery file")

embed_if_html('other/input/2021_wqx_data/summer_2021_wqx_data/SGS/SGS_results_Summer2021.pdf', text = "Download Original Summer 2021 Metals/Nutrients Lab Results from SGS - PDF file")

```

\**Note: the chain of custody documents for SGS Laboratories are integrated into the above downloadable PDF files.*

\newpage

```{r, 2021 WQX formatting for SGS, echo = F, message = F, warning = F}


# clear environment
rm(list=ls())

# restore embed_if_html helper after environment clear
source("functions/embed_if_html.R")

# load packages
library(tidyverse)
library(readxl)
library(openxlsx)
library(data.table)
library(stringr)
library(magrittr)
library(janitor)
library(hms)
library(lubridate)
library(anytime)

xfun::pkg_load2(c("htmltools", "mime"))


# ── Year Configuration ────────────────────────────────────────────────────────
# Change these values to adapt this pipeline to a different data year.
year <- 2021

# Field sampling dates for SGS/ALS/TSS ingestion (mdy() format)
spring_sample_date <- "5/11/2021"
summer_sample_date <- "7/27/2021"

# Lab receipt dates used in FC and TSS ingestion (ymd() format)
spring_rec_date <- "2021-05-11"
summer_rec_date <- "2021-07-27"

# FC lab analysis dates: the date FC samples were processed at the lab
# (typically the day after field collection). Only used in FC ingestion.
spring_fc_analysis_date <- mdy("5/12/2021")
summer_fc_analysis_date <- mdy("7/28/2021")

# FC lab source for each season. Determines which format parser runs in
# ingest_fc.R. Options: "SWWTP", "Taurianen".
# Note: Taurianen was used for summer FC in 2021 and some prior years.
#       Taurianen closed after 2021; 2022+ seasons use "SWWTP" for both.
spring_fc_lab <- "SWWTP"
summer_fc_lab <- "Taurianen"   # change to "SWWTP" for 2022+

# Explicit FC file paths (not derived from year — filenames embed sampling
# dates and vary by lab, making pattern construction unreliable).
spring_fc_path <- "other/input/2021_wqx_data/spring_2021_wqx_data/SWWTP/KRWF Fecal 05-11-21.xls"
summer_fc_path <- "other/input/2021_wqx_data/summer_2021_wqx_data/Taurianen/Fecal_Coliform_Results_Spreadsheet.xlsx"

# Explicit TSS file paths (same reason as FC).
spring_tss_path <- "other/input/2021_wqx_data/spring_2021_wqx_data/SWWTP/KRWF TSS MONITORING 05-11-21.xlsx"
summer_tss_path <- "other/input/2021_wqx_data/summer_2021_wqx_data/SWWTP/KRWF TSS MONITORING 07-28-21.xlsx"

# Derived directory paths for SGS/ALS data (consistent naming pattern).
data_dir        <- file.path("other/input", paste0(year, "_wqx_data"))
spring_data_dir <- file.path(data_dir, paste0("spring_", year, "_wqx_data"))
summer_data_dir <- file.path(data_dir, paste0("summer_", year, "_wqx_data"))
output_qaqc_dir <- file.path("other/output/lab_qaqc_data",
                              paste0(year, "_lab_qaqc_data"))

# Derived intermediate output paths
wqx_intermediate_path <- file.path("other/output/wqx_formatted/intermediate",
                                    paste0(year, "_kwf_baseline_results_wqx.csv"))
flagged_export_path   <- file.path("other/output/wqx_formatted/intermediate",
                                    paste0(year, "_export_data_flagged.csv"))
flag_decisions_path   <- file.path("other/input/wqx_templates",
                                    paste0(year, "_data_flag_decisions.csv"))
```

```{r ingest-sgs-als, echo = F, message = F, warning = F, include = F}

# Parts A–E: SGS/ALS ingestion, site-name join, analytical method join,
# non-detect handling, and lab QA/QC segregation.
# Requires: year, spring_data_dir, summer_data_dir, output_qaqc_dir (from year-config)
# Produces: dat (field results ready for FC/TSS bind)
source("functions/appendix_a_scripts/ingest_sgs_als.R")
```

<br>

##### 2021 Fecal Coliform Lab Results (Soldotna Wastewater Treatment Plant (SWWTP)/Taurianen Engineering)

```{r, echo = F}
embed_if_html('other/input/2021_wqx_data/spring_2021_wqx_data/SWWTP/KRWF Fecal 05-11-21.xls', text = "Download Original Spring 2021 Fecal Coliform Lab Results from SWWTP")

```

```{r, echo = F}
embed_if_html('other/input/2021_wqx_data/summer_2021_wqx_data/Taurianen/FecalColiform_Results_Summer2021.pdf', text = "Download Original Summer 2021 Fecal Coliform Lab Results from Taurianen")

```

```{r, echo = F}
embed_if_html('other/input/2021_wqx_data/spring_2021_wqx_data/SWWTP/SWWTP_Spring_2021_TCC_FC.jpg', text = "Download Spring 2021 Fecal Coliform Chain of Custody")
```

```{r, echo = F}
embed_if_html('other/input/2021_wqx_data/summer_2021_wqx_data/Taurianen/FC_CoC_Taurianen_Summer2021.pdf', text = "Download Summer 2021 Fecal Coliform Chain of Custody")
```

```{r ingest-fc, echo = F, message = F, warning = F, include = F}

# Spring and summer FC ingestion. Lab format determined by spring_fc_lab / summer_fc_lab.
# Requires: dat, spring/summer_fc_path, spring/summer_fc_lab, dates, output_qaqc_dir
# Produces: dat (with FC rows added)
source("functions/appendix_a_scripts/ingest_fc.R")
```

<br>

##### 2021 Total Suspended Solids Lab Results (Soldotna Wastewater Treatment Plant (SWWTP))

```{r, echo = F}
embed_if_html(spring_tss_path, text = "Download Original Spring 2021 Total Suspended Solids Results from SWWTP.xlsx")
```

```{r, echo = F}
embed_if_html(summer_tss_path, text = "Download Original Summer 2021 Total Suspended Solids Results from SWWTP.xlsx")
```

```{r, echo = F}
embed_if_html('other/input/2021_wqx_data/spring_2021_wqx_data/SWWTP/SWWTP_Spring_2021_TCC_FC.jpg', text = "Download Spring 2021 Total Suspended Solids Chain of Custody")
```

```{r, echo = F}
embed_if_html('other/input/2021_wqx_data/summer_2021_wqx_data/SWWTP/TSS_CoC_SWWTP_Summer2021.pdf', text = "Download Summer 2021 Total Suspended Solids Chain of Custody")
```

<br>

```{r ingest-tss, echo = F, message = F, warning = F, include = F}

# SWWTP TSS ingestion (spring + summer). See ingest_tss.R for QA gap notes.
# Requires: dat, spring_tss_path, summer_tss_path, spring/summer_rec_date
# Produces: dat (with TSS rows added)
source("functions/appendix_a_scripts/ingest_tss.R")
```

```{r format-wqx, echo = F, message = F, warning = F, include = F}

# Lookup joins (coords, fractions, detection conditions, preservative, container),
# WQX column construction, Activity ID build, and provisional CSV export.
# Requires: dat, year, spring_sample_date, wqx_intermediate_path
# Produces: dat (raw cols restored), dat_raw, all_dat; writes wqx_intermediate_path
source("functions/appendix_a_scripts/format_wqx.R")
```

<br>

### 2021 Provisional Results, Prior to QA/QC Review

*Results last updated `r Sys.Date()`*

The above data sources have been collated in to a single .csv file (available for download) into a format compatible with the EPA Water Quality Exchange. ***These data have not yet been evaluated against QA/QC standards following guidance in the current project Quality Assurance Project Plan.***

```{r, echo = F}
embed_if_html(wqx_intermediate_path, text = "Download All Provisional 2021 Kenai River Baseline Water Quality Monitoring Results, Formatted for EPA WQX import")

```

<br>

------------------------------------------------------------------------

### 2021 Data QA/QC Evaluation

Prior to upload to the EPA WQX, all water quality data must be checked against a series of standard questions in order to evaluate how quality assurance / quality check (QA/QC) requirements are met. The draft Data Evaluation Checklist Template (available for download below) outlines these questions:

```{r, echo = F}
embed_if_html('other/documents/wqx_documents/Kenai_Baseline_Data_Evaluation_Checklist_20230331.xlsx', text = "Download 2021 Draft Kenai Baseline Data Evaluation Checklist Template")

```

#### Pre-Database

##### Overall Project Success

Note: the sequence of questions is structured such that data will be gradually flagged at each step based on the criteria described.

**1.) Were the appropriate analytical methods used for all parameters?**

Yes. Analytical methods from the approved 2020 QAPP were employed.

<br>

**2.) Were there any deviations from the sampling plan?**

All sites were visited as planned on 5/11/2021 and 7/27/2021. Most intrinsic water quality parameters measured with instruments (pH, dissolved oxygen, conductivity, turbidity) were not measured, which is a deviation of from the QAPP.

<br>

**3.) Were field duplicates, blanks, and/or other QC samples collected as planned?**

```{r echo = F, message = F}
### First: get number of samples (results) actually collected from formatted, collated results created from the above code chunks
### report for each parameter and date
total_samples_collected_2021_summary_param <- read.csv(wqx_intermediate_path) %>%
  clean_names() %>%
  group_by(result_analytical_method_id,characteristic_name,activity_start_date,activity_type) %>%
  count() %>%
  rename(actual_results_n = n) %>%
  transform(activity_start_date = ymd(activity_start_date))



### Second: use kit request data to count "required" (planned) samples

## read in SPRING 2021 planned samples from the packing list
spring_planned <- read_excel(file.path(spring_data_dir, "SGS", paste0("KRBWQM_SGS_bottle_order_Spring_", year, ".xlsx")), sheet = "bottles_analyses_spring_2021") %>%
  mutate(activity_start_date = spring_sample_date)

## read in SUMMER 2021 planned samples from the packing list
summer_planned <- read_excel(file.path(summer_data_dir, "SGS", paste0("KRBWQM_SGS_bottle_order_Summer_", year, ".xlsx")), sheet = "bottles_analyses_summer_2021") %>%
  mutate(activity_start_date = summer_sample_date)


## combine spring and summer 2021 planned samples
planned_samples <- bind_rows(spring_planned,summer_planned) %>%
  
## prepare and summarise table samples collected and tests performed
  clean_names() %>%
  filter(item == "Bottle") %>%
  select(monitoring_location_id,site,container_size,quantity,preservative,analysis_1,analysis_2,activity_start_date) %>%
  fill(site, .direction = "down") %>%
  fill(monitoring_location_id, .direction = "down") %>%
  filter(site != "Spare Kit") %>%
  # make sample type names consistent
  mutate(sample_type = case_when(
    grepl("DUPLICATE",site, ignore.case = TRUE) ~ "FIELD DUPLICATE",
    grepl("blank",analysis_1, ignore.case = TRUE) ~ "TRIP BLANK"
  )) %>%
  pivot_longer(values_to = "analysis", cols = c("analysis_1","analysis_2")) %>%
  filter(!is.na(analysis)) %>%
  
  # consolidate BTEX analysis types ("blank" analysis is same as analysis regular samples; difference is only that no sample was collected for the trip blanks)
  
  # distinction for trip blanks (for BTEX) listed in "activity type" column
  
  mutate(analysis = str_replace(analysis,"624 - BTEX trip blank","624 - BTEX")) %>%
  
  # rename analysis type names to match between planned and actual 
  mutate(analysis = case_when(
    analysis == "SM4500-NO3E - Nitrogen (Nitrate+Nitrite)" ~ "4500-NO3(F)",  
    analysis == "SM4500-PE - Total Phosphorus" ~ "4500-P-E",
    analysis == "200.7 - Total Metals" ~ "200.7",
    analysis == "200.8 - Dissolved Metals" ~ "200.8",
    analysis == "1 liter Total Suspended Solids" ~ "2540-D",
    analysis == "Fecal Coliform" ~ "9222D",
    analysis == "624 - BTEX" ~ "8260D")) %>%
  
  # assign number of expected result types per analysis
  mutate(expected_results = case_when(
    analysis == "4500-NO3(F)" ~ 1, # Nitrate/Nitrite                           
    analysis == "4500-P-E" ~ 1, # total phosphorus
    analysis == "200.7" ~ 3,          # total metals
    analysis == "200.8" ~ 6, # custom short list; note that lab accidentally ran full list of 27 analytes in 2021
    analysis == "2540-D" ~ 1, # TSS
    analysis == "9222D" ~ 1,
    analysis == "8260D" ~ 6)) %>% # BTEX analyses
  mutate(activity_type = case_when(
    sample_type == "FIELD DUPLICATE" ~ "Field Duplicate",
    sample_type == "TRIP BLANK" ~ "Trip Blank",
    TRUE ~ "Field Msr/Obs")) %>%
  select(-sample_type)

# clear intermediary dataframes steps
# rm(spring_planned,summer_planned)


# create summary table of PLANNED samples
planned_samples_summary <- planned_samples %>%
  group_by(analysis,expected_results,activity_start_date,activity_type) %>%
  summarise(expected_ct = sum(expected_results)) %>%
  ungroup() %>%
  select(-expected_results) %>%
  rename(expected_results_n = expected_ct,
         result_analytical_method_id = analysis) %>%
  transform(activity_start_date = mdy(activity_start_date),
            expected_results_n = as.numeric(expected_results_n)) %>%
  # replace naming conventions in "activity_type" column to match join dataframe
  mutate(activity_type = case_when(
    activity_type == "Field Duplicate" ~ "Quality Control Field Replicate Msr/Obs",
    activity_type == "Field Msr/Obs" ~ "Field Msr/Obs",
    activity_type == "Trip Blank" ~ "Quality Control Sample-Trip Blank"
  ))

# BY SITE
# create summary table of PLANNED samples BY SITE
planned_samples_summary_by_site <- planned_samples %>%
  group_by(monitoring_location_id,analysis,expected_results,activity_start_date,activity_type) %>%
  summarise(expected_ct = sum(expected_results)) %>%
  ungroup() %>%
  select(-expected_results) %>%
  rename(expected_results_n = expected_ct,
         result_analytical_method_id = analysis) %>%
  transform(activity_start_date = mdy(activity_start_date),
            expected_results_n = as.numeric(expected_results_n)) %>%
  # replace naming conventions in "activity_type" column to match join dataframe
  mutate(activity_type = case_when(
    activity_type == "Field Duplicate" ~ "Quality Control Field Replicate Msr/Obs",
    activity_type == "Field Msr/Obs" ~ "Field Msr/Obs",
    activity_type == "Trip Blank" ~ "Quality Control Sample-Trip Blank"
  ))

  
### create table of quantity of samples (results) actually collected; not grouped by parameter, but instead by analysis type
total_samples_collected_2021_summary <- read.csv(wqx_intermediate_path) %>%
  clean_names() %>%
  group_by(result_analytical_method_id,activity_start_date,activity_type) %>%
  count() %>%
  rename(actual_results_n = n) %>%
  transform(activity_start_date = ymd(activity_start_date)) 


# create table of planned vs actual ANALYSES
planned_actual_analyses_2021 <- left_join(planned_samples_summary,total_samples_collected_2021_summary,
                                          by = c("result_analytical_method_id",
                                                 "activity_start_date",
                                                 "activity_type")) %>%
  mutate(pct_diff = ((actual_results_n - expected_results_n) / (actual_results_n + expected_results_n)) * 100)

# write table to download
write.csv(planned_actual_analyses_2021,"other/output/field_qa_qc_data/planned_actual_analyses_2021.csv", row.names = F)



```

```{r, echo = F}
embed_if_html('other/output/field_qa_qc_data/planned_actual_analyses_2021.csv', text = "Download Planned vs. Actual Analysis Results for 2021, Kenai Baseline")

```

<br>

To see a table comparing planned vs actual results for 2021, download and view the csv file linked above.

From the above table we can see that there are deviations between planned results and actual results available. These reasons for the deviations are known and are attributable to two causes:

***Cause 1:*** The Spring 2021 Chain of Custody (COC) from KWF to SGS was completed erroneously. The COC specified for EPA Method 200.8 analyses (dissolved metals) to be complete for all sites (when they should have stopped upstream of Morgan's Landing RM31), and it also specified for EPA Method 200.7 analyses (total metals) to stop upstream of Morgan's Landing (when they should have been performed for all sites in the project).

As a result, for Spring 2021 total metals data will be unavailable for sites upstream of the Morgan's Landing RM31 site.

***Cause 2:*** For Summer 2021, the SGS laboratory performed the EPA Method 200.8 analyses (dissolved metals) for all 27 analytes available for the method; instead of just the smaller subset of analytes as requested. (E.g., KWF received extra data for free. In this case., there are no consequences of deviating from the planned analyses).

<br>

**4.) Do the laboratory reports provide results for all sites and parameters?**

The laboratory reports provide results for all sites, and for all parameters, with the exceptions outlined above in question #3.

<br>

**5.) Is a copy of the Chain of Custody included with the laboratory reports?**

We worked with three separate laboratories in 2021:

-   SGS Laboratories, Anchorage, AK

    -   Chain of Custody documents are included within the PDF laboratory reports linked above earlier in this appendix.

-   Soldotna Wastewater Treatment Plant, Soldotna, AK

    -   Chain of Custody documents are on file with Kenai Watershed Forum for fecal coliform and total suspended solids for 5/11/2021, and for total suspended solids on 7/27/2021.

-   Tauriainen Engineering & Testing, Soldotna, AK

    -   An individual document for each sample reports the time and date of delivery and analysis for each sample. These documents are included with the PDF laboratory reports linked above earlier in this appendix.

<br>

**6.) Do the laboratory reports match the Chain of Custody and requested methods throughout?**

The laboratory reports match the Chain of Custody and requested methods, with the one exception discussed in question #3.

For summer 2021, the SGS laboratory performed the EPA Method 200.8 analyses for all 27 analytes available for the method; instead of just the smaller subset of analytes as requested. (E.g., KWF received extra data for free. In this case., there are no consequences of deviating from the planned analyses).

<br>

**7.) Are the number of samples on the laboratory reports the same as on the Chain of Custody?**

The quantity of sample bottles sent to the laboratories matches the number of analyzed samples for samples collected and delivered on 5/11/2021 and 7/27/2021.

<br>

**8.) Was all supporting info provided in the laboratory report, such as reporting limits for all analyses and definitions?**

We worked with three separate laboratories in 2021:

-   SGS Laboratories, Anchorage, AK

    -   SGS provided data as PDFs which included reporting limits, as well as in the form of an Electronic Data Deliverable where this information is also included in column format.

-   Soldotna Wastewater Treatment Plant (SWWTP), Soldotna, AK

    -   SWWTP provided data in the form of .xls files. Reporting limits are described in this project's current Quality Assurance Action Plan.

-   Tauriainen Engineering & Testing, Soldotna, AK

    -   Tauriainen provided data in the form of PDF documents. Reporting limits are described in this project's current Quality Assurance Action Plan.

<br>

**9.) Are site names, dates, and times correct and as expected?**

-   *Yes, after post-season correction documented in this report.* Notes: In 2021 Kenai Watershed Forum used pre-printed waterproof labels on all sample bottles, reducing opportunity for field and lab transcription errors. Remaining site name transcription errors from laboratories were corrected in post-season data review, using scripted code that did not modify the original report.

<br>

**10.) Were there any issues with instrument calibration?**

-   Instruments to measure intrinsic water quality parameters (sondes; to measure pH, dissolved oxygen, conductivity, turbidity) were not employed in 2021.

-   Teams did use hand-held probes to record water temperature on-site. Prior to field use, the hand-held probes were verified as measuring within the accuracy level defined in the QAPP using an ice bath in the Kenai Watershed Forum laboratory.

<br>

**11.) Did the instruments perform as expected?**

-   The hand-held water temperature probes performed as expected in 2021.

<br>

**12.) Was instrument calibration performed according to the QAPP and instrument recommendations?**

-   Water temperature is a parameter that is "verified" rather than calibrated. The hand-held water temperature probes were verified as measuring within the accuracy level defined in the QAPP using an ice bath in the laboratory, according to instrument recommendations.

<br>

**13.) Was instrument verification during the field season performed according to the QAPP and instrument recommendations?**

-   The hand-held water temperature probes were verified as measuring within the accuracy level define in the QAPP using an ice bath in the laboratory, according to instrument recommendations.

<br>

**14.) Were instrument calibration verification logs or records kept?**

-   Yes. These records are held at Kenai Watershed Forum, 44129 Sterling Hwy, Soldotna, AK.

<br>

**15.) Do the instrument data files site IDs, time stamps and file names match?**

-   Instrument files were not employed in 2021. Measurements from hand held probes were recorded on waterproof paper field forms.

<br>

**16.) Is any insitu field data rejected and why?**

-   No insitu data is rejected from 5/11/2021 or 7/27/2021.

<br>

**17.) Were preservation, hold time and temperature requirements met?**

-   Yes. Summer and Spring 2021 holding time requirements were met for all samples. See downloadable files below. Laboratory result documents indicated no compromises of preservation and temperature requirements.

```{r hold-times, echo = F, message = F, warning = F}

# embed for download
embed_if_html("other/input/wqx_templates/sample_holding_times.csv", text = "Download Table of Maximum Holding Times for Each Sample Type")
```

```{r hold-times-1, echo = F, message = F, warning = F}

# define columns to retain
col_names <- c(
  "sample",
  "epa_analysis_id",
  "analyte",
  "collect_date",
  "collect_time",
  "rec_date",
  "rec_time",
  "lab_name")

# read in data
total_samples_collected_2021_summary_param <- read.csv(wqx_intermediate_path)

# read in columns relevant to holding time calculations
holdtime_dat <- dat %>%
  select(all_of(col_names)) %>%
  # some observations have date/time in separate columns, while some have them in the same column. remedy this inconsistency
  mutate(rec_date1 = as.Date(rec_date),
         rec_time1 = case_when(
           !is.na(rec_time) ~ as_hms(rec_time),
           is.na(rec_time) ~ as_hms(rec_date))) %>%
  select(-rec_time,-rec_date) %>%
  rename(rec_time = rec_time1,
         rec_date = rec_date1) %>%
  # create single date/time columns for activity and lab receipt
  mutate(activity_datetime = ymd_hms(paste(collect_date,collect_time)),
         rec_datetime = ymd_hms(paste(rec_date,rec_time))) %>%
  # calculate actual holding time period in hours
  mutate(hold_time_hours = as.numeric(rec_datetime - activity_datetime)) %>%
  select(-collect_date,-collect_time,-rec_date,-rec_time) %>%
  mutate(analytical_method = str_trim(dat$analytical_method, side = "left"))
  
  
# join the (manually created) dataframe with maximum allowed sample holding times
# read in max holding times
max_holding_times <- read.csv("other/input/wqx_templates/sample_holding_times.csv") %>%
  transform(max_holding_time_hours = as.numeric(max_holding_time_hours)) 

# join holding time data to maximum holding times table
# calculate Y/N column for pass/fail
holdtime_dat <- left_join(holdtime_dat,max_holding_times) %>%
  mutate(hold_time_pass = case_when(
    hold_time_hours > max_holding_time_hours ~ "N",
    TRUE ~ "Y"))
```

```{r echo = F}
# save hold time calculation results as external csv
write.csv(holdtime_dat,"other/output/field_qa_qc_data/holding_time_calcs.csv", row.names = F)

# embed for download
embed_if_html("other/output/field_qa_qc_data/holding_time_calcs.csv", text = "Download Holding Time Calculations for Spring and Summer 2021 Field Samples")

```

<br>

**18.) Are dissolved metal quantities less than total metals quantities?**

-   Dissolved metals results are ***not*** consistently less than total metals results, as would be logically anticipated (see post from [Flowlink Environmental](https://www.flowlink.ca/post/2018/12/15/dissolved-total-metals-whats-the-difference-and-why-it-matters) website for description and rationale):

    -   [**"Dissolved metal concentration** is determined by filtering a water sample through 0.45 uM filter. Water that passes through the filter is analyzed for metals and the result is reported as dissolved metal concentration. **Total metal concentration** is determined by analyzing the unfiltered sample.]{.underline}

        [**"Total metal concentration = Dissolved metal concentration + Particulate (insoluble) metal concentration.**]{.underline}"

    -   In 2021, there are results available for both dissolved and total metals for three elements: Ca, Fe, and Mg; for sites at and downstream of RM 31.

    -   Dissolved metals results are available for 7/27/2021 (summer) only, and not for 5/11/2021 (spring); see question #3 in this sequence.

```{r echo = F, message = F, warning = F}

# define elements to compare
diss_total_analytes <- c("Calcium, Total","Iron, Total","Magnesium, Total",
                         "Calcium","Iron","Magnesium")


# prep dataframe
# All Ca/Fe/Mg values in 'dat' are in mg/L after the unit correction applied at ingestion.
# Units are standardized to mg/L here to match the CDX upload dataset.
diss_total_metals_dat <- dat %>%
  select(analyte, collect_date, collect_time, monitoring_location_name, result, units, result_detection_condition, sample_condition) %>%
  filter(analyte %in% diss_total_analytes) %>%
  mutate(result = case_when(
    units == "mg/l" ~ result,         # normalize lowercase label
    units == "ug/L" ~ result / 1000,  # convert any residual ug/L to mg/L
    TRUE ~ result),
    units = "mg/L")

# choose location to save csvs
dir <- "other/output/field_qa_qc_data/metals_total_diss/"


# calcium
ca <- diss_total_metals_dat %>%
  filter(analyte %in% c("Calcium","Calcium, Total")) %>%
  pivot_wider(names_from = "analyte", values_from = "result") %>%
  mutate(particulate_conc = `Calcium, Total` - Calcium) %>%
  # mutate column to designate if total metals > dissolved metals
  mutate(total_greater_diss = ifelse(particulate_conc > 0,"Y","N"))
ca_dir <- paste0(dir,"ca",".csv")
write.csv(ca,ca_dir)
embed_if_html(paste0(dir,"ca",".csv"), text = "Download 2021 Calcium data, total vs dissolved")

# iron
fe <- diss_total_metals_dat %>%
  filter(analyte %in% c("Iron","Iron, Total")) %>%
  pivot_wider(names_from = "analyte", values_from = "result") %>%
  mutate(particulate_conc = `Iron, Total` - Iron) %>%
  # mutate column to designate if total metals > dissolved metals
  mutate(total_greater_diss = ifelse(particulate_conc > 0,"Y","N"))
fe_dir <- paste0(dir,"fe",".csv")
write.csv(fe,fe_dir)
embed_if_html(paste0(dir,"fe",".csv"), text = "Download 2021 Iron data, total vs dissolved")

# magnesium
mg <- diss_total_metals_dat %>%
  filter(analyte %in% c("Magnesium","Magnesium, Total")) %>%
  pivot_wider(names_from = "analyte", values_from = "result") %>%
  mutate(particulate_conc = `Magnesium, Total` - Magnesium) %>%
  # mutate column to designate if total metals > dissolved metals
  mutate(total_greater_diss = ifelse(particulate_conc > 0,"Y","N"))
mg_dir <- paste0(dir,"mg",".csv")
write.csv(mg,mg_dir)
embed_if_html(paste0(dir,"mg",".csv"), text = "Download 2021 Magnesium data, total vs dissolved")

# --- Summary scalars for metals comparison prose (computed before objects are removed) ---

# Calcium: pct difference = (dissolved - total) / total * 100 (positive = dissolved > total)
ca <- ca |> mutate(
  pct_diff = (Calcium - `Calcium, Total`) / `Calcium, Total` * 100,
  abs_diff = abs(Calcium - `Calcium, Total`)
)
ca_paired     <- ca |> filter(!is.na(Calcium) & !is.na(`Calcium, Total`))
ca_n_sites    <- nrow(ca_paired)
ca_pct_min    <- paste0(round(min(ca_paired$pct_diff),   1), "%")
ca_pct_max    <- paste0(round(max(ca_paired$pct_diff),   1), "%")
ca_pct_median <- paste0(round(median(ca_paired$pct_diff), 0), "%")
ca_abs_min    <- round(min(ca_paired$abs_diff), 1)
ca_abs_max    <- round(max(ca_paired$abs_diff), 1)

# Iron: all paired sites expected to show total > dissolved
fe_n_sites <- sum(!is.na(fe$Iron) & !is.na(fe$`Iron, Total`))

# Magnesium
mg <- mg |> mutate(pct_diff = (Magnesium - `Magnesium, Total`) / `Magnesium, Total` * 100)
mg_paired      <- mg |> filter(!is.na(Magnesium) & !is.na(`Magnesium, Total`))
mg_n_sites     <- n_distinct(mg_paired$monitoring_location_name)
mg_n_expected  <- sum(mg_paired$total_greater_diss == "Y", na.rm = TRUE)

# RM 0 NNC: most extreme case where dissolved > total
mg_rm0_pct <- mg_paired |>
  filter(grepl("River Mile 0", monitoring_location_name), pct_diff > 0) |>
  slice_max(pct_diff, n = 1) |>
  mutate(label = paste0("~", round(pct_diff, 1), "%")) |>
  pull(label)

# RM 1.5
mg_rm15_row <- mg_paired |> filter(grepl("River Mile 1.5", monitoring_location_name))
mg_rm15_pct  <- paste0("~", round(mg_rm15_row$pct_diff[1], 1), "%")
mg_rm15_conc <- paste0("~", round(mg_rm15_row$Magnesium[1], 0), " mg/L")

rm(ca, fe, mg)

```

Results:

-   **Calcium:** All `r ca_n_sites` summer 2021 sites with paired dissolved and total measurements show dissolved \> total. The dissolved Ca concentration exceeds total Ca by `r ca_pct_min` to `r ca_pct_max` (median \~`r ca_pct_median`) across all sites. This is a systematic, consistent pattern that is too large and too uniform to be attributable to measurement uncertainty alone.

-   **Iron:** All `r fe_n_sites` summer 2021 sites with paired data show dissolved \< total, which is the expected relationship. Iron behaves as anticipated.

-   **Magnesium:** `r mg_n_expected` of `r mg_n_sites` summer 2021 sites show total \> dissolved, as expected. Two sites near the Kenai River mouth are exceptions: RM 0 NNC (dissolved exceeds total by `r mg_rm0_pct`) and RM 1.5 (dissolved exceeds total by `r mg_rm15_pct`). These small percentage differences are likely within the typical analytical uncertainty for ICP methods (\~5–10%) and may not represent a meaningful discrepancy. Note that the Mg concentration at RM 1.5 is anomalously high (`r mg_rm15_conc`, compared to 1–8 mg/L at other lower-river sites), possibly reflecting brackish/tidal influence near Cook Inlet.

<br>

The calcium discrepancy is unresolved. Dissolved Ca consistently exceeds total Ca by approximately `r ca_pct_min` to `r ca_pct_max` at all `r ca_n_sites` paired summer 2021 sites. This pattern is too systematic and too large (`r ca_abs_min`–`r ca_abs_max` mg/L absolute difference) to be explained by measurement uncertainty alone.

Two possible explanations are worth investigating in future years:

1.  **Inter-laboratory bias.** Dissolved Ca (and Mg, Fe) were analyzed by SGS Laboratories using EPA Method 200.8, while total Ca (and Mg, Fe) were analyzed by ALS Laboratories using EPA Method 200.7. A consistent offset between two different labs analyzing the same matrix via different methods could produce a systematic directional bias for one or more analytes.

2.  **Method-specific matrix effects.** EPA Method 200.7 (ICP-OES) and 200.8 (ICP-MS) differ in their digestion and detection approaches, and inter-method differences for specific analytes in fresh water matrices have been documented in the literature. Calcium in particular may be subject to spectral interferences or calibration differences between the two methods.

It is also worth noting that the summer 2021 SGS dissolved metals data (200.8) was affected by a unit labeling error in the original EDD (values reported with "ug/L" labels but at mg/L concentrations), which was corrected at data ingestion. While the corrected values are internally consistent and physically plausible, the original data anomaly introduces additional uncertainty about the SGS 200.8 dissolved metals results specifically.

The magnesium exceptions at RM 0 NNC (`r mg_rm0_pct`) and RM 1.5 (`r mg_rm15_pct`) are small enough to fall within typical ICP analytical uncertainty and are not considered significant anomalies. Iron behaves as expected at all sites.

**This discrepancy for calcium warrants targeted follow-up.** Future years should include split samples sent to both labs for direct inter-laboratory comparison, or use a single lab for both dissolved and total fractions, to determine whether the offset is reproducible and attributable to a specific cause. Until resolved, the 2021 Ca dissolved and total results should be interpreted with caution relative to each other.

<br>

**19.) Are the duplicate sample(s) RPD (Relative Percent Difference) within range described in QAPP?**

In 2021, field duplicate samples were collected at two sites on both 5/11/2021 and 7/27/2021:

-   No Name Creek (Tributary, RM0)
-   Funny River (Tributary, RM30)

Following the example in the Alaska Department of Environmental Conservation 2021 Kenai River Metals Field Report [@apsens2022]: "... a set of paired samples was evaluated for RPD only if:

-   a.) one or both of the samples were above the Limit of Quantitation (LOQ); AND if
-   b.) one or both of the samples were at least two times the LOQ."

```{r echo = F, message = F, warning = F}

# generate table of rpd values and make csv downloadable

# general approach:
# create table with RPD goals from QAPP
# calculate observed 2021 RPD values
# compare to goals

# subset field data from sites with field duplicates in 2021

# for 2021, we have the same sites for both dates. In future years we will need to assign sites by dates as well (join by date).

qaqc_sites_2021 <- read_excel("other/input/wqx_templates/wqx_qaqc/wqx_qaqc_info.xlsx", sheet = "field_dup_sites_2021") %>%
  remove_empty()

# read in data formatted for export
export_dat <- read.csv(wqx_intermediate_path) %>%
  clean_names()

# inner join
field_dup_dat_2021 <- inner_join(export_dat,qaqc_sites_2021) 

# specify and retain columns needed for RPD analysis

rpd_cols_1 <- c("monitoring_location_id",
              "activity_type",
              "activity_start_date",
              "characteristic_name",
              "result_detection_condition",
              "result_value",
              "result_unit",
              "result_detection_limit_type_1",
              "result_detection_limit_value_1",
              "result_detection_limit_unit_1"
             )

# prepare dataframe to evaluate if pairs of values can be evaluated for RPD
rpd_check_dat <- field_dup_dat_2021 %>%
  select(all_of(rpd_cols_1)) %>%
  pivot_wider(names_from = "activity_type", values_from = c("result_value")) %>%
  clean_names() %>%
  
# assign columns to check if data can be evaluated for RPD

# field samples
  
# is field sample above LOQ?
  mutate(field_above_loq = case_when(
    field_msr_obs > result_detection_limit_value_1 ~ "Y",
    TRUE ~ "N")) %>%
# is field sample 2x LOQ ?
  mutate(field_loq_2x = case_when(
    field_msr_obs > 2*result_detection_limit_value_1 ~ "Y",
    TRUE ~ "N")) %>%
  
# qc samples
  
# is qc sample above LOQ?
  mutate(qc_above_loq = case_when(
    quality_control_field_replicate_msr_obs > result_detection_limit_value_1 ~ "Y",
    TRUE ~ "N")) %>%
# is qc sample 2x LOQ ?
  mutate(qc_loq_2x = case_when(
    quality_control_field_replicate_msr_obs > 2*result_detection_limit_value_1 ~ "Y",
    TRUE ~ "N"))

```

```{r echo = F}
# execute 2-stage winnowing process to determine which data can be used for RPD calculations

# evaluate if observations can be use for RPD calculations
rpd_check_dat %<>%
  
  # does this pair of measurements meet criteria that allow them to be used for RPD calculations?
  # 1.) one or both of the measurements must be above the parameter's limit of quantitation (LOQ), AND
  # 2.) one or both of the measurements must be at least two times the LOQ."
  
  mutate(loq_pass = case_when(
    field_above_loq == "Y" | qc_above_loq == "Y" ~ "Y",
    TRUE ~ "N")) %>%
  mutate(loq_2x_pass = case_when(
    field_loq_2x == "Y" | qc_loq_2x == "Y" ~ "Y",
    TRUE ~ "N")) %>%
  # pair of values is eligible for RPD calcs if above criteria met, AND both values are present
  mutate(rpd_eligible = case_when(
    loq_pass == "Y" & loq_2x_pass == "Y" & !is.na(field_msr_obs) & !is.na(quality_control_field_replicate_msr_obs) ~ "Y" ,
         TRUE ~ "N"))
  
# calculate RPD values
rpd_check_dat %<>%

  # calc absolute percent diff between field and qc sample value
  mutate(rpd_pct = ((field_msr_obs - quality_control_field_replicate_msr_obs) / 
                      ((field_msr_obs + quality_control_field_replicate_msr_obs)/2))*100) %>%
  mutate(rpd_pct = abs(round(rpd_pct,2))) 

# apply result flags where appropriate according to RPD logic described below
#z <- rpd_check_dat %>%
#  mutate(resultFlag = case_when(
#    field
#  ))
  
# check 2016 report to see how this was executed exactly .... seems like a ton of data 

# write csv
write.csv(rpd_check_dat,"other/output/field_qa_qc_data/rpd_check_dat.csv", row.names = F)

# --- Summary scalars for RPD narrative prose ---
rpd_n_total    <- nrow(rpd_check_dat)
rpd_n_eligible <- sum(rpd_check_dat$rpd_eligible == "Y", na.rm = TRUE)

# Apply QAPP thresholds: Phosphorus/Nitrates = 25%, Suspended solids = 5%, others = 20%
rpd_eligible_dat <- rpd_check_dat |>
  filter(rpd_eligible == "Y") |>
  mutate(threshold = case_when(
    grepl("Nitrate|Nitrite",  characteristic_name, ignore.case = TRUE) ~ 25,
    grepl("Phosphorus",       characteristic_name, ignore.case = TRUE) ~ 25,
    grepl("suspended|solids", characteristic_name, ignore.case = TRUE) ~ 5,
    TRUE ~ 20
  )) |>
  filter(!is.na(rpd_pct))

rpd_exceed     <- rpd_eligible_dat |> filter(rpd_pct > threshold)
rpd_n_exceed   <- nrow(rpd_exceed)
rpd_exceed_tbl <- rpd_exceed |> count(characteristic_name) |> arrange(desc(n))

# Fecal coliform RPD counts
rpd_fc_eligible <- rpd_eligible_dat |>
  filter(grepl("Fecal|Coliform", characteristic_name, ignore.case = TRUE))
rpd_fc_n_total  <- nrow(rpd_fc_eligible)
rpd_fc_n_high   <- sum(rpd_fc_eligible$rpd_pct > 60, na.rm = TRUE)
```

A csv file of Relative Percent Difference values organized by site, date, and parameter may be downloaded below:

```{r echo = F}
# export csv of rpd results 
embed_if_html("other/output/field_qa_qc_data/rpd_check_dat.csv", text = "Download 2021 Duplicate Sample Relative Percent Difference Data")

# note: code in the next chunk references an object named "rpd_check_dat," do not change the name of the csv

#z <- rpd_check_dat %>%
#  select(monitoring_location_id,characteristic_name,activity_start_date,rpd_pct) %>%
#  arrange(characteristic_name)
```

<br>

From the above downloadable table, we can see that a limited subset of paired values from field duplicates are eligible to be used for RPD calculations for 2021. Pairs of measurements must meet LOQ criteria described above ("a" and "b") in order to be used. For 2021 data, of `r rpd_n_total` paired field duplicate observations, `r rpd_n_eligible` of them meet these criteria and thus may be used for RPD calculations.

RPD values range from `r min(rpd_check_dat$rpd_pct, na.rm = T)` % to `r max(rpd_check_dat$rpd_pct, na.rm = T)` %.

In the 2020 Quality Assurance Action Plan, Relative Percent Difference values for most parameters are set at 20%, with the exception of "Phosphorus, total" and "Nitrates (NO2+NO3)" at 25%, and "Suspended solids, total" at 5%.

Using the available RPD values against the above criteria, `r rpd_n_exceed` paired observations exceed project precision goals in 2021. The parameters with exceedances are: `r paste(paste0(rpd_exceed_tbl$characteristic_name, " (n=", rpd_exceed_tbl$n, ")"), collapse = ", ")`.

Most other parameters have at least one available RPD value that is within project precision goals defined in the QAPP.

The parameters that do *not* have at least one RPD calculation available, due to the LOQ constraints described above, are the following:

-   Arsenic
-   Cadmium
-   Chromium
-   Copper
-   Lead
-   Zinc

*Other notes*

In 2021 the full suite of dissolved metals was run for method 200.8 on samples from 7/27/2021. However we intend to only submit data for parameters that had a pre-defined QA plan as outlined in the QAPP, which is the case for the following dissolved six metals: Mg, As, Cd, Cr, Cu, Pb, and Zn.

**Summary of Discussion on RPD (Relative Percent Difference) Values for 2021**

We attempted to calculate RPD values for all parameters collected in 2021 at two field sites on two dates. For each pair of duplicate observations, one of the following descriptions apply:

-   It ***was possible*** to calculate RPD values because:
    -   a.) one or both of the samples **were above** the LOQ; AND one or both of the samples **were** at least two times the LOQ
        -   Project data for this parameter will be submitted unflagged to EPA WQX if RPD values meet project goals outlined in the QAPP
-   It ***was not possible*** to calculate RPD values because:
    -   a.) one or both of the samples **were not above** the LOQ; OR one or both of the samples **were not** at least two times the LOQ
        -   Project data for this parameter will be submitted unflagged to EPA WQX if both paired values are non-detect

KWF's approach to flagging when RPD cannot be calculated follows the example of the Alaska Department of Environmental Conservation (ADEC) 2021 Kenai River Metals Field Report (Apsens and Petitt, 2021): when duplicate pairs cannot be evaluated for RPD because concentrations are at or below the LOQ, results are retained as accepted non-detect or estimated values rather than flagged. A below-LOQ result is not a precision failure; it reflects actual environmental conditions. Non-detect status is already documented in the `Result Detection Condition` column (U = not detected; J = detected below LOQ, treated as an estimate). KWF communicated with the Alaska Department of Environmental Conservation on August 22, 2023 via an inquiry on this topic. The memo includes a visual flow chart of the logic for when data is flagged, and is available for download below.

```{r echo = F}
# export csv of rpd results 
embed_if_html("other/documents/QAPP/memos/KWF Memo to ADEC August 2023.pdf", text = "Download August 2023 Memo to ADEC regarding RPD Interpretation")

```

<br>

*Final determinations on flagging / not flagging 2021 data submitted to the EPA WQX*

***All 2021 field results are submitted unflagged to the EPA WQX unless discussed below.***

-   *Fecal Coliform*

    -   Given that `r rpd_fc_n_high` out of `r rpd_fc_n_total` RPD values for fecal coliform are \>60%, field data values for the 2021 dataset may vary from actual environmental conditions. The 2020 KWF QAPP does not specify an RPD value for fecal coliform, it only specifies control checks for sterility and temperature. We used EPA Method 9222D (drinking water) for fecal coliform analysis methods, which does not specify an RPD value. However EPA general recommendations for pathogens in sewage sludge recommends and RPD of 20% [(link)](https://www.epa.gov/biosolids/quality-assurance-project-plan-qapp).

    -   Given the above considerations, KWF has submitted 2021 fecal coliform data to the EPA WQX as "flagged" (Result Status ID = Rejected). Fecal coliform data is evaluated in this report and assessed against state standards, but given the high RPD values among duplicate samples, 2021 data should be interpreted cautiously in the context of potential regulatory exceedances.

    -   A comparison of fecal coliform RPD values across years (2014–2022) and against results from the 2016 comprehensive report is recommended as a future step.

-   *Selected Dissolved Metals (As, Cd, Cr, Cu, Pb, Zn)*

    -   RPD calculations could not be performed for these parameters because concentrations at both the primary and duplicate sample locations were at or below the LOQ. Following the ADEC 2021 approach described above, these results are submitted to EPA WQX as **Accepted**, with non-detect or estimated status clearly documented in the `Result Detection Condition` and `Result Qualifier` columns. This approach recognizes that a below-LOQ result is not a precision failure.
    -   Note: beginning in Summer 2022, KWF initiated a series of changes to the QAPP that will better facilitate QA/QC decisions regarding dissolved metals results in future years. These changes include use of a dissolved metals field blank and changing from field filtration to lab filtration as standard practice, both of which will improve the ability to detect contamination and calculate RPD for dissolved metals.

<br>

**20.) Were there any laboratory discrepancies, errors, data qualifiers, or QC failures (review laboratory duplicates, matrix spikes and blanks)?**

Laboratory discrepancies and QC failures from the SGS lab are described on page 2 of the spring and summer PDF document reports, which are available for download below.

```{r echo = F}
embed_if_html('other/input/2021_wqx_data/spring_2021_wqx_data/SGS/Spring 2021 Results from SGS.pdf', text = "Download Original Spring 2021 Metals/Nutrients Lab Results from SGS - PDF file")

embed_if_html('other/input/2021_wqx_data/spring_2021_wqx_data/SGS/Spring 2021 Results from SGS.pdf', text = "Download Original Spring 2021 Metals/Nutrients Lab Results from SGS - PDF file")

```

```{r echo = F, message = F, warning = F}
# calculate qa/qc discrepancies as detailed in PDF reports check. qapp for guidance, or check w/ qa officer...

matrix_spike_limits <- read.csv(file.path(output_qaqc_dir, "sgs_als_qaqc_dat.csv")) %>%
  select(lab_name,sample,collect_date,extracted_date,sample_type,result,analyte,dissolved,analytical_method,
         resultflag,percent_recovered,rec_limit_low,rec_limit_high,sample_rpd,rpd_limit_low,rpd_limit_high,
         loq,lod,detection_limit,sample_condition) %>%
  
  # Matrix Spike Recovery
  # calculate if matrix spike recovery limits pass requirements
  mutate(rec_limit_pass = case_when(
    is.na(rec_limit_low) ~ "",
    percent_recovered > rec_limit_high |
      percent_recovered < rec_limit_low ~ "N",
    TRUE ~ "Y"
  ))

# calculate total number of matrix spike cases where recovery exceeded limits
matrix_spike_recovery_fails <- matrix_spike_limits %>%
  filter(rec_limit_pass == "N") %>%
  nrow() %>%
  as.numeric()

# generate table of specific failed samples
matrix_spike_recovery_fails_tbl <- matrix_spike_limits %>%
  filter(rec_limit_pass == "N") %>%
  select(-extracted_date,-dissolved)
write.csv(matrix_spike_recovery_fails_tbl,file.path("other/output/lab_qaqc_data", paste0("matrix_spike_recovery_fails_", year, ".csv")))

# --- Summary scalars for matrix spike prose ---
ms_recovery_min <- round(min(matrix_spike_recovery_fails_tbl$percent_recovered), 0)
ms_recovery_max <- round(max(matrix_spike_recovery_fails_tbl$percent_recovered), 0)
ms_bias_min     <- ms_recovery_min - 100
ms_bias_max     <- ms_recovery_max - 100

```

<br>

The following discrepancies were identified in the results from SGS laboratories:

**Matrix Spike Recoveries**

A total of `r matrix_spike_recovery_fails` matrix spike or matrix spike duplicate samples are outside of QC criteria. The limit of recovery range for the analyte "Total Nitrate/Nitrite-N" is 90% - 110%. For these matrix spike measurements outside of the QC criteria, recovery levels range from `r ms_recovery_min`% to `r ms_recovery_max`%. The subset of sample values that exceed threshold may be downloaded as a table below:

```{r matrix-spikes, echo = F, messages = F, warnings = F}
# embed for download
embed_if_html(file.path("other/output/lab_qaqc_data", paste0("matrix_spike_recovery_fails_", year, ".csv")), text = "Download Matrix Spike Recovery Values Exceeding Threshold for 2021 Sample Analyses")

```

<br>

No additional laboratory QA/QC anomalies for any other parameters were noted by any laboratories contracted in 2021. Additional details are available upon request at [hydrology\@kenaiwatershed.org](mailto:hydrology@kenaiwatershed.org){.email}

As of `r Sys.Date()`, "Total Nitrate/Nitrite-N" values will be flagged for the spring 2021 dataset. (Spring only and not summer because no matrix spike recovery exedences were observed from the summer 2021 lab results). The matrix spike recovery exceedances indicate that reported lab values may be `r ms_bias_min`% to `r ms_bias_max`% above actual field values. At a later date we will evaluate these results in the context of year 2000 - present data to help evaluate the relevancy of this flag.

```{r echo = F}

# apply "flag = Y" to all "Total Nitrate/Nitrite-N" values for Spring 2021

# this step is already accomplished in a previous code block, which imports and joins flag_decisions_path to the parameters chosen in this discussion.

```

<br>

**21.) Is any laboratory data rejected and why?**

No additional 2021 laboratory data currently indicates a need for being flagged or rejected.

<br>

**22.) During the field season, review raw data files (EDDs \[Electronic Data Deliverables\], instrument records) as they are received. Document changes and corrections to methods as needed.**

In 2021, KWF initiated as standard practice the receipt of machine-readable EDD files from contracted laboratories, in addition to PDF results. The EDD files are being reviewed as part of this QA/QC process, which was initiated in Winter 2022.

<br>

**23.) Is the dataset complete and did you receive the expected number of results?**

From the discussion and results in question #3, deviations are evident between planned and actual results reported. The reasons for the deviations are known and are attributable to the two factors discussed in question #3, and thus not concerning to project integrity.

<br>

**24.) Was the data collected representative of environmental conditions?.**

The data collected in this project is representative of seasonal hydrological conditions on two fieldwork days in 2021. Conditions vary instantaneously thus this dataset serves to provide representative longitudinal snapshots of the watershed across space on the two dates.

Flagged data indicates that lab results may not be representative of environmental conditions, and thus may require additional interpretation or exclusion from analysis.

<br>

**25.) Does project meet Completeness Measure A criteria?**

From the QAPP, Completeness Measure A: "the primary number of samples collected divided by the usable number of samples submitted to ADEC, with a goal of 85% completeness." (Note that as of 2023 results are now submitted directly to EPA rather than DEC, per request of DEC).

(Note: metals parameters not specified in the QAPP are excluded from these analyses. In 2021, SGS laboratories provided dissolved metals results for parameters beyond what was originally requested.)

We calculated Completeness Measure A at the following project scales:

-   For individual parameters
-   For individual sites
-   As a project whole

```{r apply-flags, echo = F, message = F, warning = F}

# Join flag decisions and write flagged export CSV (must run before Q25/Q26).
# Requires: wqx_intermediate_path, flag_decisions_path, flagged_export_path
# Produces: export_dat (in memory for downstream QA/QC chunks)
source("functions/appendix_a_scripts/apply_qaqc_flags.R")
```

```{r echo = F, message=FALSE}
# read in list of analytes specified in QAPP
analytes_qapp <- read.csv("other/input/wqx_templates/analytes_list_manual_edit.csv")
colnames(analytes_qapp) <- c("characteristic_name","a","param_in_qapp") 
analytes_qapp %<>% select(-a)

# BY PARAMETER
# summarise flags by parameter
completeness_summary_param <- read.csv(flagged_export_path) %>%
  clean_names() %>%
  # reduce the list of analytes to only those specified in QAPP
  left_join(analytes_qapp) %>%
  filter(param_in_qapp == "Y") %>%
  # summarise
  group_by(result_analytical_method_id,characteristic_name,flag) %>%
  count() %>%
  rename(actual_results_n = n) %>%
 # transform(activity_start_date = ymd(activity_start_date)) 
  pivot_wider(names_from = "flag", values_from = "actual_results_n") %>%
  rename(flag_Y = Y,
         flag_N = N) %>%
  rowwise() %>%
  mutate(total_samples = sum(flag_Y,flag_N, na.rm = T)) %>%
  # calculate Completeness Measure A for each parameter
  mutate(cma = flag_N/total_samples)%>%
  mutate(cma = replace_na(cma,0)) %>%
  adorn_pct_formatting(,,,cma)
write.csv(completeness_summary_param,"other/output/field_qa_qc_data/completeness_measures/cma_parameter.csv")


## BY SITE
# summarise flags by site
completeness_summary_site <- read.csv(flagged_export_path) %>%
  clean_names() %>%
  # reduce the list of analytes to only those specified in QAPP
  left_join(analytes_qapp) %>%
  filter(param_in_qapp == "Y") %>%
  # summarise
  group_by(monitoring_location_id,flag) %>%
  count() %>%
  rename(actual_results_n = n) %>%
 # transform(activity_start_date = ymd(activity_start_date)) 
  pivot_wider(names_from = "flag", values_from = "actual_results_n") %>%
  rename(flag_Y = Y,
         flag_N = N) %>%
  rowwise() %>%
  mutate(total_samples = sum(flag_Y,flag_N, na.rm = T)) %>%
  # calculate Completeness Measure A for each parameter
  mutate(cma = flag_N/total_samples)%>%
  mutate(cma = replace_na(cma,0)) %>%
  adorn_pct_formatting(,,,cma)
write.csv(completeness_summary_site,"other/output/field_qa_qc_data/completeness_measures/cma_site.csv")


# OVERALL
# summarise flags for overall project
completeness_summary_project <- read.csv(flagged_export_path) %>%
  clean_names() %>%
  # reduce the list of analytes to only those specified in QAPP
  left_join(analytes_qapp) %>%
  filter(param_in_qapp == "Y") %>%
  # summarise
  group_by(flag) %>%
  count() %>%
  rename(actual_results_n = n) %>%
 # transform(activity_start_date = ymd(activity_start_date)) 
  pivot_wider(names_from = "flag", values_from = "actual_results_n") %>%
  rename(flag_Y = Y,
         flag_N = N) %>%
  rowwise() %>%
  mutate(total_samples = sum(flag_Y,flag_N, na.rm = T)) %>%
  # calculate Completeness Measure A for each parameter
  mutate(cma = flag_N/total_samples)%>%
  mutate(cma = replace_na(cma,0)) %>%
  adorn_pct_formatting(,,,cma)
write.csv(completeness_summary_project,"other/output/field_qa_qc_data/completeness_measures/cma_project.csv")


# procedure: based on discussion in above questions, decide what is flagged as "unusable." THEN return to this question to calculate completeness measure A. adapt from existing code in "appendix_a_temp.qmd"


```

```{r complete-measures, echo = F, messages = F, warnings = F}
# embed for download
embed_if_html("other/output/field_qa_qc_data/completeness_measures/cma_parameter.csv", text = "Download Completeness Measure A Results, By Parameter")
embed_if_html("other/output/field_qa_qc_data/completeness_measures/cma_site.csv", text = "Download Completeness Measure A Results, By Site")
embed_if_html("other/output/field_qa_qc_data/completeness_measures/cma_project.csv", text = "Download Completeness Measure A Results, By Overall")

```

<br>

```{r cma-narrative-values, echo = F, message = F}
# Compute scalar values used in the CMA discussion prose below
cma_dat_inline <- read.csv(flagged_export_path) |>
  clean_names() |>
  left_join(analytes_qapp) |>
  filter(param_in_qapp == "Y")

cma_total        <- nrow(cma_dat_inline)
cma_unflagged_n  <- sum(cma_dat_inline$flag == "N", na.rm = TRUE)
cma_flagged_n    <- sum(cma_dat_inline$flag == "Y", na.rm = TRUE)
cma_overall_pct  <- paste0(round(cma_unflagged_n / cma_total * 100, 1), "%")
cma_flagged_pct  <- paste0(round(cma_flagged_n   / cma_total * 100, 1), "%")

cma_site_vals <- cma_dat_inline |>
  group_by(monitoring_location_id) |>
  summarise(cma_val = sum(flag == "N") / n(), .groups = "drop")
cma_site_min <- paste0(round(min(cma_site_vals$cma_val) * 100, 1), "%")
cma_site_max <- paste0(round(max(cma_site_vals$cma_val) * 100, 1), "%")
```

*Discussion on "Completeness Measure A" Results*

Based on the data flagging decisions discussed in the previous questions, the 2021 dataset achieved Completeness Measure A goals for most parameters.

When completeness results were summarized by parameter, two methods had 0% Completeness Measure A: Fecal Coliform (EPA Method 9222D, both seasons) and Total Nitrate/Nitrite-N (SM 4500-NO3(F), spring 2021 only). All other in-QAPP parameters achieved 100% CMA. Notably, dissolved metals (EPA Method 200.8) achieved 100% CMA - RPD calculations could not be performed because concentrations were at or below the LOQ, but per KWF's adopted approach (see Q19), below-LOQ results are retained as Accepted rather than flagged.

When completeness results were summarized by site, Completeness Measure A values ranged from `r cma_site_min` to `r cma_site_max`, depending on the site's sampling plan.

When completeness results are summarized for the overall 2021 dataset, Completeness Measure A is calculated at `r cma_overall_pct`; indicating that `r cma_flagged_pct` of the overall result values were flagged.

<br>

**26.) Does project meet Completeness Measure B criteria?**

From the QAPP, Completeness Measure B: "the planned number of samples divided by the useable samples (submitted to ADEC) with a goal of 60%."

For these calculations, we will consider "usable" as "unflagged," and will calculate at the following scales:

-   For individual parameters (analysis methods)
-   For individual sites
-   As a project whole

```{r echo = F, message=FALSE, eval= T}

# NOTE: IF THE NUMBER/TYPE OF PLANNED SAMPLES IS IDENTICAL TO THE SAMPLES ACTUALLY COLLECTED,
# THEN COMPLETENESS MEASURES A = COMPLETENESS MEASURES B

# BY PARAMETER (by analytical method)
# unflagged (usable) vs planned results
cmb_by_method <- read.csv(flagged_export_path) %>%
  clean_names() %>%
  # reduce the list of analytes to only those specified in QAPP
  left_join(analytes_qapp) %>%
  filter(param_in_qapp == "Y") %>%
  group_by(result_analytical_method_id, activity_start_date, activity_type, flag) %>%
  count() %>%
  rename(unflagged_results_n = n) %>%
  pivot_wider(names_from = "flag", values_from = "unflagged_results_n") %>%
  rename(flag_Y = Y,
         flag_N = N) %>%
  transform(activity_start_date = ymd(activity_start_date)) %>%
  left_join(planned_samples_summary) %>%
  # CMB = unflagged / planned (per QAPP formula T - (I + NC) / T, goal 60%)
  mutate(cmb = flag_N / expected_results_n) %>%
  mutate(cmb = replace_na(cmb, 0)) %>%
  adorn_pct_formatting(,,,cmb)
write.csv(cmb_by_method, "other/output/field_qa_qc_data/completeness_measures/cmb_parameter.csv")


# BY SITE
# Aggregate planned totals to one row per site (summing across methods and sampling events)
# so the join is 1-to-1 on monitoring_location_id
planned_by_site_totals <- planned_samples_summary_by_site %>%
  group_by(monitoring_location_id) %>%
  summarise(expected_results_n = sum(expected_results_n), .groups = "drop")

cmb_by_site <- read.csv(flagged_export_path) %>%
  clean_names() %>%
  # reduce the list of analytes to only those specified in QAPP
  left_join(analytes_qapp) %>%
  filter(param_in_qapp == "Y") %>%
  group_by(monitoring_location_id, flag) %>%
  count() %>%
  rename(unflagged_results_n = n) %>%
  pivot_wider(names_from = "flag", values_from = "unflagged_results_n") %>%
  rename(flag_Y = Y,
         flag_N = N) %>%
  left_join(planned_by_site_totals) %>%
  # CMB = unflagged / planned
  mutate(cmb = flag_N / expected_results_n) %>%
  mutate(cmb = replace_na(cmb, 0)) %>%
  adorn_pct_formatting(,,,cmb)
write.csv(cmb_by_site, "other/output/field_qa_qc_data/completeness_measures/cmb_site.csv")


# BY PROJECT WHOLE
# Sum unflagged results across entire project, divide by total planned
total_planned_2021 <- sum(planned_samples_summary$expected_results_n)

cmb_project <- read.csv(flagged_export_path) %>%
  clean_names() %>%
  left_join(analytes_qapp) %>%
  filter(param_in_qapp == "Y") %>%
  group_by(flag) %>%
  count() %>%
  pivot_wider(names_from = "flag", values_from = "n") %>%
  rename(flag_Y = Y,
         flag_N = N) %>%
  mutate(total_planned = total_planned_2021,
         cmb = flag_N / total_planned) %>%
  mutate(cmb = replace_na(cmb, 0)) %>%
  adorn_pct_formatting(,,,cmb)
write.csv(cmb_project, "other/output/field_qa_qc_data/completeness_measures/cmb_project.csv")

```

```{r echo = F, message=FALSE}
embed_if_html("other/output/field_qa_qc_data/completeness_measures/cmb_parameter.csv", text = "Download Completeness Measure B Results, By Parameter")
embed_if_html("other/output/field_qa_qc_data/completeness_measures/cmb_site.csv", text = "Download Completeness Measure B Results, By Site")
embed_if_html("other/output/field_qa_qc_data/completeness_measures/cmb_project.csv", text = "Download Completeness Measure B Results, Overall")
```

```{r cmb-narrative-values, echo = F, message = F}
# Compute scalar values used in the CMB discussion prose below
cmb_usable      <- cmb_project$flag_N
cmb_planned_n   <- cmb_project$total_planned
cmb_overall_pct <- paste0(round(cmb_usable / cmb_planned_n * 100, 1), "%")

cmb_spring_totmet_pct <- cmb_by_method |>
  filter(result_analytical_method_id == "200.7",
         as.character(activity_start_date) == "2021-05-11",
         activity_type == "Field Msr/Obs") |>
  pull(cmb)
```

*Discussion on "Completeness Measure B" Results*

The project-level CMB for 2021 is **`r cmb_overall_pct`** (`r cmb_usable` usable results out of `r cmb_planned_n` planned), which is above the QAPP goal of 60%. Two parameters contributed to the CMB falling short of 100%:

1.  **Fecal coliform (EPA Method 9222D): 0% CMB, both seasons.** RPD values exceeded the 20% threshold for multiple duplicate pairs, and all 2021 fecal coliform results were designated Rejected.

2.  **Spring 2021 Total Nitrate/Nitrite-N (SM 4500-NO3(F)): 0% CMB.** Matrix spike recovery exceedances in the spring 2021 SGS lab report led to all spring nitrate results being designated Rejected. No exceedances were observed in summer 2021, which achieved 100% CMB.

Parameters achieving 100% CMB include: dissolved metals (EPA Method 200.8, both seasons - all results were at or below the LOQ and are retained as Accepted per KWF's adopted approach; see Q19), total metals (EPA Method 200.7, summer only - spring was affected by the chain-of-custody error described in Q1), TSS (SM 2540-D), Total Phosphorus (SM 4500-P-E), summer nitrate/nitrite, and BTEX (EPA Method 8260D, summer only). Spring 2021 total metals achieved `r cmb_spring_totmet_pct` CMB, reflecting the reduced site coverage resulting from the spring 2021 chain-of-custody error.

<br>

**27.) Was the QA officer consulted for any data concerns?**

*Note: the 2012 and 2019 approved Quality Assurance Action Plan (QAPP) documents list Alaska Dept of Environmental Conservation Staff as the project's QA officer* [@kenaiwatershedforum2012; @kenaiwatershedforum2019]*, whereas the 2023 approved QAPP lists Kenai Watershed Forum staff as the project QA officer* [@kenaiwatershedforum2023]*.*

Kenai Watershed Forum consulted regularly with staff from the Alaska Department of Environmental Conservation and EPA Region 10 Office in order to prepare 2021 field data and confirm field methods and data QA/QC procedures.

<br>

**28.) Are the correct monitoring locations associated with the project? Are latitude and longitude filled out in a consistent format?**

We used the leaflet package to generate a map of unique sites from the 2021 dataset:

```{r echo = F, message = F, warning = F}

# read in locations from exported data (export_dat)
export_dat <- read.csv(flagged_export_path)

site_check <- export_dat %>%
  distinct(monitoring_location_id,activity_latitude,activity_longitude) %>%
  transform(activity_longitude = as.numeric(iconv(activity_longitude, 'utf-8', 'ascii', sub=''))) %>%

  # the row for KR RM 43 longitude likely contains a non-unicode character, thus the "iconv" function, otherwise it becomes NA. The above transform also leave that row as latitude > 0, thus the "case_when" step below. Sheesh!
  
  mutate(activity_longitude = case_when(
    monitoring_location_id == "KR RM 43" ~ activity_longitude*-1,
    TRUE ~ activity_longitude
  ))
  

# generate map
library(leaflet)

if (knitr::is_html_output()) {
  leaflet(site_check) %>%
    addTiles() %>%
    addMarkers(~activity_longitude,~activity_latitude,
               popup = ~as.character(monitoring_location_id),
               label = ~as.character(monitoring_location_id))
} else {
  knitr::kable(site_check, caption = "2021 monitoring locations")
}

```

In the above map we can see that coordinates for the 2021 field sites correspond to the locations listed in the Quality Assurance Project Plan, thus they are filled out in a consistent format.

<br>

**29.) Are the QAPP and other supporting documents attached?**

The QAPP will be included and attached in the data submission to the EPA WQX. Find the current approved QAPP posted on Kenai Watershed Forum's website at <https://www.kenaiwatershed.org/news-media/qapp-revisions-completed-2023/>.

<br>

**30.) Is all project metadata correct?**

Project metadata has been reviewed and confirmed correct prior to CDX submission. The project description, monitoring location, and results/activities files were each prepared in code (see the upload section below) and verified against the existing WQX project record. The organization ID (`Kenai_WQX`), project ID, site coordinates, and QAPP approval status are all confirmed correct.

<br>

**31.) Is the organization ID correct?**

Kenai Watershed Forum's Organization ID in the EPA Water Quality Exchange is "Kenai_WQX." This value is manually confirmed during the upload process.

<br>

**32.) Are the time zones consistent and correct (AKDT in summer)?**

```{r echo = F}

# Existing format does not include AKDT time zone column.

# read in locations from exported data (export_dat)
export_dat <- read.csv(flagged_export_path) %>%
  mutate(time_zone = "AKDT")
# re-write export_dat
write.csv(export_dat,flagged_export_path)

```

All times zones are listed as AKDT.

<br>

**33.) Are all media types included? Media types appropriate to Characteristic?**

"Water" is the only media type considered in this project. All "Characteristics" (parameters) are measured from water samples or in-situ water conditions.

<br>

**34.) Check Sample Collection, Preparation and Preservation Methods, Thermal Preservative, Equipment ID, Activity Media. Is supporting information included and correct?**

```{r echo = F}
# sample collection method
unique(export_dat$equipment_id)

```

"Sample collection method": all sample values have "Equipment ID" listed as "Water Bottle."

"Preparation":this information corresponds to the "result sample fraction" column

"Preservation Methods": this information corresponds to the "preservative" and "thermal preservative" columns.

<br>

**35.) Are all expected activity types present and are QC samples correctly identified?**

```{r echo = F}
# id unique
unique(export_dat$activity_type)

```

<br>

The activity types present are listed above. All three types were planned and anticipated. QC samples are described earlier in Question #3.

<br>

**36.) Is the Activity media subdivision filled in (if relevant)?**

```{r echo = F}

unique(export_dat$activity_media_subdivision_name)

```

All "Activity Media Subdivision" names are listed as "Surface Water."

<br>

**37.) For Water activity media, is the relative depth filled in?**

Yes, "activity depth height measure" and "activity depth height measure unit" are included as columns.

<br>

**38.) Is the number of results for each Characteristic correct?**

Questions #2 and #3 address this topic in detail. A summary table of the total number of observations for each characteristic is available in the downloadable table below.

```{r echo = F}
#summarise
chr_num <- export_dat %>%
  group_by(characteristic_name) %>%
  tally()

# save
dir <- file.path("other/output/misc", paste0("characteristic_count_", year, ".csv"))
write.csv(chr_num,dir)
```

```{r echo = F}
# embed for download
embed_if_html(dir, text = "Download Observation Count by Characteristic, 2021 Kenai River Baseline Water Quality")
```

<br>

**39.) Do the range of result values make sense?**

Prior to upload to the EPA WQX, the range of results will be evaluated visually and numerically in each characteristic-specific section of the comprehensive report, relative to past results. Likely outliers will be identified and recorded in a dedicated spreadsheet (download below). Each likely outlier will have accompanying notes discussing the choice, and a decisions as to whether a more detailed statistical investigation is warranted.

```{r echo = F}
embed_file("other/input/outliers/baseline_outliers.xlsx", text = "Download identified potential outliers for 2021 Kenai River Baseline Water Quality data")

```

<br>

**40.) Are units correct and consistent for each parameter?**

```{r echo = F}

z <- export_dat %>%
  select(characteristic_name, result_unit) %>%
  distinct()
data.frame(unique(z$result_unit))

# apply correction to make units consistent. May be different solution here for other years

export_dat[export_dat == "mg/l"] <- 'mg/L'

```

We observe that equivalent units, "mg/L" and "mg/l", have slightly different notation among parameters. We will correct this at this step to make terms consistent.

<br>

**41.) Are detection limits and laboratory qualifiers included for analytical results?**

Laboratory qualifiers including "J" (below limit of quantitation; is an estimate), "=", (observed value), "U" (below detection limit), and "ND" (Not Detected) are included in the column titled "result_qualifier."

Detection limits and their corresponding units are included in columns titled "result_detection_limit_type_1" and in "result_detection_limit_unit_1" corresponding to Limit of Quanititation, and in columns titled "result_detection_limit_type_2" and "result_detection_limit_unit_2" corresponding to Method Detection Level.

<br>

**42.) Are results trip blanks and/or field blanks above detection limits, and is any related data flagged as a result?**

In summer 2021, trip blanks were included with those field crews tasked with collection water samples for hydrocarbon analyses. Sites visited by these crews include RM 1.5 - Kenai City Dock, RM 6.5 - Cunningham Park, Rm 40 - Bing's Landing, and RM 43 - Upstream of Dow Island.

On the summer 2021 SGS chain of custody documents, BTEX trip blanks are specified for the above sites. However in the reported results, a single sample named "Trip Blank" is reported. This error resulted from a mis-communication with SGS staff.

In this case, all hydrocarbon results in the Trip Blank are reported as below detection limit, suggesting there is not a need to flag summer 2021 hydrocarbon results.

<br>

### Uploading 2021 data to the EPA Water Quality Exchange (EPA WQX)

Each water quality dataset that is housed in the EPA WQX consists of three uploads:

-   Project description

-   Monitoring locations

-   Results and activities

We will prepare and finalize each of these uploads here.

#### Project Description

On 12/5/2023 we downloaded the KWF Agency Baseline project description datafile from the EPA Water Quality Portal ([EPA WQP](https://www.waterqualitydata.us/#countrycode=US&statecode=US%3A02&countycode=US%3A02%3A122&within=65&lat=60.550309&long=-151.098437&organization=KENAI_WQX&mimeType=csv&providers=NWIS&providers=STEWARDS&providers=STORET)). The file can be directly accessed at the link below:

```{r echo = F}

embed_file("other/input/WQX_downloads/project/project.csv", text = "Kenai River Baseline Water Quality Monitoring project details for EPA WQP")

```

<br>

We found that the project file listed the QAPP (Quality Assurance Project Plan) as "not yet approved." It is possible that this designation is listed as such because previously approved versions of the QAPP (versions 1.0 and 2.0) do not include signatures from an EPA staff on the Title & Approval page, but were instead approved by Alaska Department of Environment Conservation Staff. For the [QAPP approved on 5/9/2023](https://www.kenaiwatershed.org/news-media/qapp-revisions-completed-2023/) (version 3.0), two Region 10 EPA staff signed as approvers on the Title & Approval page.

To apply this correction, in the "Project" spreadsheet we modified "QAPP Approval Status" from "N" to "Y" and re-uploaded to the EPA Water Quality Exchange in Winter 2023.

#### Monitoring locations

On 12/5/2023 we downloaded the KWF Agency Baseline project monitoring locations datafile from the EPA Water Quality Portal ([EPA WQP](https://www.waterqualitydata.us/#advanced=true)). The file can be directly accessed at the link below:

```{r echo = F}

embed_file("other/input/WQX_downloads/station/station.csv", text = "Download Kenai River Baseline Water Quality Monitoring Sites 2021")

```

<br>

We retained all sites whose column description for "MonitoringLocationName" contained the prefix "KBL" (Kenai Baseline).

We plotted locations on a map using the leaflet package, and confirmed that existing site names in the database matched those used in the 2021 dataset.

In the EPA WQX Central Data Exchange, we created a custom import configuration to accommodate importing the list of existing sites.

```{r echo = F, message=FALSE}

# import csv of sites downloaded from EPA Water Quality Portal
wqp_sites <- read.csv("other/input/WQX_downloads/station/station.csv")

# read in locations from exported data (export_dat)
wqp_sites %<>%
  distinct(MonitoringLocationName,MonitoringLocationDescriptionText,LatitudeMeasure,LongitudeMeasure) %>%
  filter(str_detect(MonitoringLocationName,"^KBL"))
  
# generate map
if (knitr::is_html_output()) {
  leaflet(wqp_sites) %>%
    addTiles() %>%
    addMarkers(~LongitudeMeasure,~LatitudeMeasure,
               popup = ~as.character(MonitoringLocationName),
               label = ~as.character(MonitoringLocationName))
} else {
  knitr::kable(wqp_sites, caption = "KBL monitoring locations from EPA WQP")
}

# Notes on ADEC Site Names
# It appears that the "monitoring location id" site names provided by ADEC are not in the existing project dataset on the EPA WQX. (Example: ADEC template calls the No Name Creek site "KR RM 0 NNC", whereas the existing name in the EPA database for this site is "KBL_t_00.0", indicating "Kenai River Baseline, tributary, river mile 0.00")

# For the next step, "Results & Activities," we need to use the existing EPA site names for MonitoringLocationID, and also integrate these into the Activity ID names, and also make sure other relevant columns like MonitoringLocationDescriptionText are included.

# we can perform the join by lat/long if we convert to character and have the same number of decimal places in both dataframes
# this issue will be avoidable in the future!

# Notes on "station.csv" download from EPA WQX
# The download contains extensive column info about well characteristics. No wells are included as part of this study, but the full column set is retained for consistency with the CDX import configuration.


```

<br>

```{r echo = F, message = F, warning = F}

# Flagging was applied at Q25 above via source("functions/appendix_a_scripts/apply_qaqc_flags.R").
# The flagged export CSV (flagged_export_path) is available in memory as export_dat.
```

<br>

#### Results and Activities

Within Kenai Watershed Forum's EPA Water Quality Portal web account, we developed an import configuration that accommodated all relevant columns and data types for this project. We based the configuration primarily on the template provided by the Alaska Dept of Environmental Conservation.

For this export, we identified as "Rejected" all observations that had been "flagged" throughout the QA/QC process described above. Flagged observations will be uploaded to the EPA WQX labeled as "Rejected" under the column "Result Status ID."

```{r generate-cdx-export, echo = F, message = F, warning = F}

# Apply Result Status ID, finalize Results & Activities format, and write
# results_activities.csv, project.csv, station.csv for CDX upload.
# Requires: export_dat (with time_zone applied in Q32 chunk above)
source("functions/appendix_a_scripts/generate_cdx_export.R")
```

**Notes for 2021 CDX upload:**

-   The 2021 dataset includes 13 "Phosphorus" results from EPA Method 200.8 (Dissolved Metals by ICP/MS), in addition to 48 "Total Phosphorus" results from SM21 4500-P-E. The Method 200.8 phosphorus results are excluded from the CDX submission because this method was not part of the QAPP. Only the `Total Phosphorus` / `4500-P-E` results are uploaded. The filter is applied in the "Results and Activities" code block above.

-   Corrections to data from years other than 2021 (e.g., spring 2013 specific conductance units) are documented in AGENTS.md and will be addressed in a dedicated chapter of the kenai-river-wqx-qaqc repo, not here.

\newpage
